---
title: "Clinical Regression Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clinical Regression Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
```

This tutorial shows the unified regression interface in `leo.ukb`:

- `leo_cox_regression()` for time-to-event outcomes
- `leo_logistic_regression()` for binary outcomes
- `leo_linear_regression()` for continuous outcomes

Across all three functions, the design is intentionally consistent:

- one input data frame
- one exposure at a time
- one outcome at a time
- one named model list carried into the formatted output
- default paper-style wide output in `$result`
- optional tidy output through the paired formatter function

```{r setup, include = FALSE}
suppressPackageStartupMessages(library(leo.ukb))
```

## Named model specification

All three regression helpers accept `x_cov` as a named list. Those names are
preserved in the formatted result table, which makes the final columns more
readable for manuscript work. Each dataset can define its own covariate names;
the key point is that the model labels stay stable across formatted outputs.

```{r}
model_list <- list(
  "Crude" = NULL,
  "Model A" = c("sex"),
  "Model B" = c("sex", "bmi_group")
)

model_list
```

## Cox regression

For Cox regression, `y_out` is a length-2 character vector:

- the first element is the event column
- the second element is the follow-up time column

```{r}
lung_df <- dplyr::transmute(
  survival::lung,
  outcome = as.integer(status == 2),
  outcome_censor = time,
  age = age,
  sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
  ecog_group = factor(
    ph.ecog,
    levels = 0:3,
    labels = c("ECOG0", "ECOG1", "ECOG2", "ECOG3")
  )
)

cox_model <- list(
  "Crude" = NULL,
  "Model A" = c("sex"),
  "Model B" = c("sex", "ecog_group")
)

res_cox <- leo_cox_regression(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_cov = cox_model,
  verbose = FALSE
)

res_cox$result
```

The default `$result` is the manuscript-style wide table. If you want a long
table for plotting or batch binding, use the paired formatter:

```{r}
dplyr::select(
  leo_cox_regression_format(res_cox, style = "tidy"),
  Model, Exposure, Outcome, HR, `95% CI`, `P value`
)
```

Categorical exposures can be handled in the same framework by setting
`x_exp_type = "categorical"` when needed:

```{r}
res_cox_cat <- leo_cox_regression(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "ecog_group",
  x_cov = list("Crude" = NULL, "Model A" = c("sex")),
  x_exp_type = "categorical",
  verbose = FALSE
)

res_cox_cat$result
```

## Logistic regression

For logistic regression, `y_out` is a single binary outcome column name and
`case_value` defines which value should be treated as the case group.

```{r}
set.seed(123)
n <- 240

logi_df <- data.frame(
  outcome = rbinom(n, 1, 0.35),
  prs = rnorm(n),
  sex = factor(sample(c("Male", "Female"), n, replace = TRUE)),
  bmi_group = factor(
    sample(c("Low", "Mid", "High"), n, replace = TRUE),
    levels = c("Low", "Mid", "High")
  )
)

res_logi <- leo_logistic_regression(
  df = logi_df,
  y_out = "outcome",
  x_exp = "prs",
  x_cov = model_list,
  case_value = 1,
  verbose = FALSE
)

res_logi$result
```

```{r}
dplyr::select(
  leo_logistic_regression_format(res_logi, style = "tidy"),
  Model, Exposure, Outcome, OR, `95% CI`, `P value`
)
```

## Linear regression

For linear regression, `y_out` is a continuous outcome column name and the rest
of the interface stays the same.

```{r}
linear_df <- dplyr::transmute(
  mtcars,
  outcome = mpg,
  wt = wt,
  sex = factor(am, levels = c(0, 1), labels = c("Auto", "Manual")),
  bmi_group = factor(cyl, levels = c(4, 6, 8), labels = c("Cyl4", "Cyl6", "Cyl8"))
)

res_linear <- leo_linear_regression(
  df = linear_df,
  y_out = "outcome",
  x_exp = "wt",
  x_cov = model_list,
  verbose = FALSE
)

res_linear$result
```

```{r}
dplyr::select(
  leo_linear_regression_format(res_linear, style = "tidy"),
  Model, Exposure, Outcome, Beta, `95% CI`, `P value`
)
```

## Notes for clinical workflows

- All models are fitted on the same complete-case cohort within each analysis
  call, so crude and adjusted results remain directly comparable.
- If rows are removed because of missing values, the functions print a warning
  and suggest `leo_impute_na()` as a follow-up option.
- `style = "gtsummary"` is supported for all three regression families when
  both `gtsummary` and `broom.helpers` are available.