---
title: "Clinical Regression Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clinical Regression Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(leo.ukb))
```

This tutorial introduces the unified regression interface in `leo.ukb`. The
goal is simple: once an analysis-ready data frame is prepared, the same grammar
should work for survival, binary, and continuous outcomes.

The current interface covers:

- `leo_cox()` for time-to-event outcomes
- `leo_logistic()` for binary outcomes
- `leo_linear()` for continuous outcomes

The Cox toolkit also extends to:

- `leo_cox_interaction()` for multiplicative interaction tests
- `leo_cox_subgroup()` for subgroup tables with `P for interaction`
- `leo_cox_mediation()` for survival mediation analysis

Across these functions, the design stays deliberately consistent:

- one analysis-ready data frame
- one exposure at a time
- one outcome at a time
- one named model list
- one shared complete-case cohort within each analysis call
- one default manuscript-style table in `$result`

## Start with a clear model list

All regression helpers accept `x_cov` as a named list. Those names are carried
forward into the final table headers, so it is worth defining the model
hierarchy carefully at the start.

```{r}
model_list <- list(
  "Crude" = NULL,
  "Model A" = c("sex"),
  "Model B" = c("sex", "bmi_group")
)

model_list
```

In practice, this means the final output will read naturally as `Crude HR`,
`Model A OR`, or `Model B Beta` without additional post-processing.

## Cox regression

For Cox models, `y_out` must be `c(event, time)`, where the first column is the
incident event indicator and the second column is the follow-up time.

We start by building a compact, analysis-ready example data frame:

```{r}
lung_df <- dplyr::transmute(
  survival::lung,
  outcome = as.integer(status == 2),
  outcome_censor = time,
  age = age,
  sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
  ecog_group = factor(ph.ecog, levels = 0:3, labels = c("ECOG0", "ECOG1", "ECOG2", "ECOG3"))
)

knitr::kable(utils::head(lung_df, 6))
```

This table is already in the shape expected by `leo_cox()`:

- `outcome`: incident event indicator
- `outcome_censor`: follow-up time
- `age`: continuous exposure in this example
- `sex` and `ecog_group`: covariates or subgroup variables

Now define the Cox model hierarchy and fit the main analysis:

```{r}
cox_model <- list(
  "Crude" = NULL,
  "Model A" = c("sex"),
  "Model B" = c("sex", "ecog_group")
)

res_cox <- leo_cox(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_cov = cox_model,
  verbose = FALSE
)

res_cox$result
```

The default output is intentionally paper-oriented: one row for the exposure,
one block of columns per model, and directly readable `HR`, `95% CI`, and
`P value`.

When the next step is plotting, row binding, or further programmatic work, the
paired formatter can return a tidy table:

```{r}
dplyr::select(
  leo_cox_format(res_cox, style = "tidy"),
  Model, Exposure, Outcome, HR, `95% CI`, `P value`
)
```

### Categorical exposure in the same framework

The interface does not change when the exposure is categorical. The only thing
that changes is that we declare `x_exp_type = "categorical"`.

```{r}
res_cox_cat <- leo_cox(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "ecog_group",
  x_cov = list("Crude" = NULL, "Model A" = c("sex")),
  x_exp_type = "categorical",
  verbose = FALSE
)

res_cox_cat$result
```

Notice that the output now contains a reference row and one row for each
exposure level. This is usually the layout needed for manuscript tables.

## Cox interaction

`leo_cox_interaction()` answers a different question from the main Cox model.
It asks whether the exposure effect differs across levels of another variable.

That is why `P for interaction` is not inferred from subgroup-specific
`P values`; it is estimated from a dedicated interaction model fitted on the
full cohort.

```{r}
res_inter <- leo_cox_interaction(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_inter = "sex",
  x_cov = list("Crude" = NULL, "Model A" = c("ecog_group")),
  verbose = FALSE
)

res_inter$result
```

In a clinical report, this table is usually interpreted alongside subgroup
tables rather than as a standalone endpoint.

## Cox subgroup

`leo_cox_subgroup()` combines two pieces of information:

- subgroup-specific HRs
- the matching `P for interaction`

By default, it can also append `P for heterogeneity` as a supplementary
summary-statistic-based metric.

```{r}
res_sub <- leo_cox_subgroup(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_subgroup = c("sex", "ecog_group"),
  x_cov = list("Crude" = NULL, "Model A" = c("ecog_group")),
  verbose = FALSE
)

utils::head(res_sub$result, 8)
```

In most applications, `P for interaction` should be treated as the primary
heterogeneity signal, with `P for heterogeneity` kept as supporting
information.

## Cox mediation

`leo_cox_mediation()` currently wraps `regmedint::regmedint()` with a Cox
survival outcome. The most reliable first-pass use cases are:

- binary exposure
- continuous mediator or binary mediator
- mostly numeric covariates

Before fitting the model, it is helpful to inspect the mediation analysis table
just as we did for Cox:

```{r}
if (requireNamespace("regmedint", quietly = TRUE)) {
  set.seed(123)
  n <- 300

  med_df <- data.frame(
    outcome = rbinom(n, 1, plogis(-0.4 + 0.7 * rbinom(n, 1, 0.5))),
    outcome_censor = rexp(n, rate = 0.08) + 0.1,
    exposure = factor(rbinom(n, 1, 0.5), levels = c(0, 1), labels = c("No", "Yes")),
    mediator = factor(rbinom(n, 1, 0.45), levels = c(0, 1), labels = c("Low", "High")),
    age = rnorm(n, 60, 8),
    bmi = rnorm(n, 26, 4)
  )

  knitr::kable(utils::head(med_df, 6))
}
```

```{r}
if (requireNamespace("regmedint", quietly = TRUE)) {
  res_med <- leo_cox_mediation(
    df = med_df,
    y_out = c("outcome", "outcome_censor"),
    x_exp = "exposure",
    x_med = "mediator",
    x_cov = list("Model A" = c("age"), "Model B" = c("age", "bmi")),
    verbose = FALSE
  )

  dplyr::select(
    res_med$result,
    Model, Effect, Estimate, `95% CI`, `P value`, `Mediator model`
  )
}
```

This output is designed for effect decomposition rather than ordinary
association reporting, so it is often best interpreted separately from the main
regression tables.

## Logistic regression

For logistic regression, `y_out` is a single binary outcome column and
`case_value` identifies the case group.

```{r}
set.seed(123)
n <- 240

logi_df <- data.frame(
  outcome = rbinom(n, 1, 0.35),
  prs = rnorm(n),
  sex = factor(sample(c("Male", "Female"), n, replace = TRUE)),
  bmi_group = factor(sample(c("Low", "Mid", "High"), n, replace = TRUE), levels = c("Low", "Mid", "High"))
)

knitr::kable(utils::head(logi_df, 6))
```

```{r}
res_logi <- leo_logistic(
  df = logi_df,
  y_out = "outcome",
  x_exp = "prs",
  x_cov = model_list,
  case_value = 1,
  verbose = FALSE
)

res_logi$result
```

```{r}
dplyr::select(
  leo_logistic_format(res_logi, style = "tidy"),
  Model, Exposure, Outcome, OR, `95% CI`, `P value`
)
```

The overall user experience is intentionally parallel to Cox: same input style,
same model list logic, same default wide table, different effect measure.

## Linear regression

For linear regression, `y_out` is continuous while the rest of the interface
stays aligned with Cox and logistic workflows.

```{r}
linear_df <- dplyr::transmute(
  mtcars,
  outcome = mpg,
  wt = wt,
  sex = factor(am, levels = c(0, 1), labels = c("Auto", "Manual")),
  bmi_group = factor(cyl, levels = c(4, 6, 8), labels = c("Cyl4", "Cyl6", "Cyl8"))
)

knitr::kable(utils::head(linear_df, 6))
```

```{r}
res_linear <- leo_linear(
  df = linear_df,
  y_out = "outcome",
  x_exp = "wt",
  x_cov = model_list,
  verbose = FALSE
)

res_linear$result
```

```{r}
dplyr::select(
  leo_linear_format(res_linear, style = "tidy"),
  Model, Exposure, Outcome, Beta, `95% CI`, `P value`
)
```

## Practical guidance

- Within a single analysis call, all models use the same complete-case cohort.
- If missingness removes rows, the functions report the row loss and suggest
  `leo_impute_na()`.
- `style = "gtsummary"` is available for Cox, logistic, and linear analyses,
  but requires both `gtsummary` and `broom.helpers`.
- For subgroup analyses, interpret `P for interaction` before
  `P for heterogeneity`.
