---
title: "Survival Analysis Step 3.1: Mediation Case Study"
description: "Apply survival mediation analysis in rare/common outcome and continuous/binary mediator case studies."
output:
  rmarkdown::html_vignette:
    highlight: default
vignette: >
  %\VignetteIndexEntry{Survival Analysis Step 3.1: Mediation Case Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(leo.ukb))
has_regmedint <- requireNamespace("regmedint", quietly = TRUE)
if (!has_regmedint) {
  knitr::opts_chunk$set(eval = FALSE)
}
fmt_num <- function(x, digits = 3) sprintf(paste0("%.", digits, "f"), x)
fmt_pct <- function(x, digits = 1) sprintf(paste0("%.", digits, "f%%"), 100 * x)
fmt_pct_ci <- function(ci_text, digits = 1) {
  ci_num <- suppressWarnings(as.numeric(trimws(strsplit(ci_text, ",", fixed = TRUE)[[1]])))
  if (length(ci_num) != 2 || anyNA(ci_num)) return(ci_text)
  paste(sprintf(paste0("%.", digits, "f%%"), 100 * ci_num), collapse = ", ")
}
get_effect_row <- function(res, code) {
  dplyr::filter(res$result, `Effect code` == code)
}
case_summary_row <- function(res, scenario) {
  te <- get_effect_row(res, "te")
  tnie <- get_effect_row(res, "tnie")
  pm <- get_effect_row(res, "pm")
  data.frame(
    Scenario = scenario,
    `Event rate` = fmt_pct(res$evaluation$event_rate[1]),
    `Outcome model` = res$evaluation$yreg[1],
    `Total effect` = sprintf("%s (%s)", fmt_num(te$Estimate[1]), te$`95% CI`[1]),
    `Indirect effect` = sprintf("%s (%s)", fmt_num(tnie$Estimate[1]), tnie$`95% CI`[1]),
    `Proportion mediated` = sprintf("%s (%s)", fmt_pct(pm$Estimate[1]), fmt_pct_ci(pm$`95% CI`[1])),
    check.names = FALSE
  )
}
```

## Why this article exists

Step 3.0 explains what mediation analysis is for. Step 3.1 is where we turn
that into a practical workflow.

The goal here is not just to run `leo_cox_mediation()`, but to answer five
questions in a way that stays tied to real output:

1. what biological question are we actually asking?
2. how do `x_exp_a0`, `x_exp_a1`, `x_med_cde`, and `x_cov_cond` map to that question?
3. when should `mreg = "linear"` be preferred over `mreg = "logistic"`?
4. what changes when `yreg = "auto"` switches from `survCox` to `survAFT_weibull`?
5. how should the scientific interpretation change across those settings?

```{r, echo = FALSE, results = "asis"}
if (!has_regmedint) {
  cat(
    "> Note: regmedint is not installed in the current build environment, ",
    "so the code below is shown but not executed. Install regmedint locally ",
    "to run the full case study.\n\n",
    sep = ""
  )
}
```

## Build one reproducible scientific story, then vary the scenario

We use a synthetic cohort with the same broad question throughout:

> is the association between higher metabolic risk and an incident outcome partly explained by inflammation?

The scenarios differ only in how we encode the mediator and how common the
outcome is.

```{r}
set.seed(20260324)

make_mediation_case <- function(n = 1500, base_log_hazard = -5.8, censor_rate = 0.06) {
  age <- rnorm(n, 60, 8)
  sex <- factor(sample(c("Female", "Male"), n, replace = TRUE))
  smoking <- factor(sample(c("Never", "Ever"), n, replace = TRUE, prob = c(0.6, 0.4)))
  sex_num <- as.integer(sex == "Male")
  smoking_num <- as.integer(smoking == "Ever")
  metabolic_risk_num <- rbinom(
    n, 1,
    plogis(-0.3 + 0.03 * (age - 60) + 0.30 * sex_num + 0.35 * smoking_num)
  )
  inflammation_score <- rnorm(
    n,
    mean = 0.85 * metabolic_risk_num + 0.04 * (age - 60) + 0.35 * smoking_num,
    sd = 0.9
  )
  hazard_rate <- exp(
    base_log_hazard +
      0.55 * metabolic_risk_num +
      0.24 * inflammation_score +
      0.02 * (age - 60) +
      0.20 * smoking_num
  )
  event_time <- rexp(n, rate = hazard_rate)
  censor_time <- rexp(n, rate = censor_rate)

  data.frame(
    incident_event = as.integer(event_time <= censor_time),
    followup_years = pmin(event_time, censor_time),
    metabolic_risk = factor(
      metabolic_risk_num,
      levels = c(0, 1),
      labels = c("Lower risk", "Higher risk")
    ),
    inflammation_score = inflammation_score,
    inflammation_group = factor(
      as.integer(inflammation_score > 0.5),
      levels = c(0, 1),
      labels = c("Lower inflammation", "Higher inflammation")
    ),
    age = age,
    sex = sex,
    smoking = smoking
  )
}

rare_case <- make_mediation_case(base_log_hazard = -5.8, censor_rate = 0.06)
common_case <- make_mediation_case(base_log_hazard = -4.6, censor_rate = 0.06)

data.frame(
  Scenario = c("Rare-outcome case", "Common-outcome case"),
  Participants = c(nrow(rare_case), nrow(common_case)),
  Events = c(sum(rare_case$incident_event), sum(common_case$incident_event)),
  `Event rate` = c(fmt_pct(mean(rare_case$incident_event)), fmt_pct(mean(common_case$incident_event)))
)
```

The rare-outcome case will let `yreg = "auto"` stay on `survCox`. The
common-outcome case is deliberately built to push the same scientific question
onto the AFT path.

## Case 1. A graded biomarker mediator in a rare-outcome setting

This is the most natural scenario when inflammation is scientifically treated
as a continuum rather than as a yes/no threshold.

```{r}
x_med_cde_case1 <- stats::median(rare_case$inflammation_score, na.rm = TRUE)

res_case1 <- leo_cox_mediation(
  df = rare_case,
  y_out = c("incident_event", "followup_years"),
  x_exp = "metabolic_risk",
  x_med = "inflammation_score",
  x_cov = c("age", "sex", "smoking"),
  x_exp_a0 = 0,
  x_exp_a1 = 1,
  x_med_cde = x_med_cde_case1,
  x_cov_cond = list(age = 60, sex = "Female", smoking = "Never"),
  mreg = "linear",
  yreg = "auto",
  interaction = TRUE,
  verbose = FALSE
)

case1_eval <- res_case1$evaluation
case1_te <- get_effect_row(res_case1, "te")
case1_tnie <- get_effect_row(res_case1, "tnie")
case1_pm <- get_effect_row(res_case1, "pm")
case1_te_ci <- case1_te[["95% CI"]][1]
case1_tnie_ci <- case1_tnie[["95% CI"]][1]
case1_pm_ci <- case1_pm[["95% CI"]][1]
```

### Read the evaluation settings before reading the effects

```{r}
res_case1$evaluation[, c(
  "exposure_contrast", "x_exp_a0", "x_exp_a1",
  "mediator_reference", "x_med_cde", "x_cov_cond",
  "mreg", "yreg", "interaction"
)]
```

This one table already answers the three user-facing evaluation questions:

- `x_exp_a0 = 0`, `x_exp_a1 = 1` means we are comparing `Lower risk -> Higher risk`
- `x_med_cde = `r fmt_num(case1_eval$x_med_cde[1])`` means the controlled direct effect is being evaluated at the median inflammation score
- `x_cov_cond = age=60; sex=Female; smoking=Never` means the reported effects are evaluated for that covariate profile, not as a profile-free population average

In plain language, the CDE here asks:

> if two otherwise similar 60-year-old never-smoking women differed only in
> metabolic-risk status, but both were fixed at an inflammation score of
> `r fmt_num(case1_eval$x_med_cde[1])`, how different would their outcome risk be?

### Visualize the biology before over-interpreting the decomposition

First look at whether the mediator really behaves like a graded biomarker across
the two exposure groups.

```{r, fig.width = 7.2, fig.height = 5.2}
ggplot2::ggplot(
  rare_case,
  ggplot2::aes(x = metabolic_risk, y = inflammation_score, fill = metabolic_risk)
) +
  ggplot2::geom_violin(trim = FALSE, alpha = 0.35, color = NA) +
  ggplot2::geom_boxplot(width = 0.16, outlier.alpha = 0.15, color = "#1f1f1f") +
  ggplot2::scale_fill_manual(values = c("#5B8FF9", "#D14A61")) +
  ggplot2::labs(
    title = "Inflammation score is shifted upward in the higher-risk group",
    x = "Metabolic risk group",
    y = "Inflammation score"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(legend.position = "none")
```

Now look at the crude joint pattern that motivates keeping `interaction = TRUE`
in the outcome model.

```{r, fig.width = 7.4, fig.height = 5.1}
joint_event_case1 <- dplyr::summarise(
  dplyr::group_by(rare_case, metabolic_risk, inflammation_group),
  n = dplyr::n(),
  event_rate = mean(incident_event),
  .groups = "drop"
)

ggplot2::ggplot(
  joint_event_case1,
  ggplot2::aes(x = inflammation_group, y = event_rate, fill = metabolic_risk)
) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.7), width = 0.62) +
  ggplot2::geom_text(
    ggplot2::aes(label = paste0(fmt_pct(event_rate), "\n(n=", n, ")")),
    position = ggplot2::position_dodge(width = 0.7),
    vjust = -0.15,
    size = 3.3
  ) +
  ggplot2::scale_fill_manual(values = c("#5B8FF9", "#D14A61")) +
  ggplot2::scale_y_continuous(
    labels = function(x) paste0(round(100 * x), "%"),
    limits = c(0, max(joint_event_case1$event_rate) * 1.18)
  ) +
  ggplot2::labs(
    title = "The highest crude event rate appears in the higher-risk plus higher-inflammation stratum",
    x = "Inflammation group",
    y = "Observed event rate",
    fill = "Metabolic risk"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

That figure does not prove interaction by itself, but it shows why the default
`interaction = TRUE` is often a sensible first pass: it avoids forcing the
effect of metabolic risk to be identical at every inflammation level.

### Interpret the actual mediation result, not an imagined one

```{r}
dplyr::filter(res_case1$result, `Effect code` %in% c("te", "tnie", "pm"))
```

Because this rare-outcome scenario stayed on `survCox`, the ratio scale is a
hazard ratio.

In this fitted example:

- the total effect was `r fmt_num(case1_te$Estimate[1])` (`r case1_te_ci`), meaning the higher-risk group had a substantially higher incident hazard overall
- the indirect pathway through the continuous inflammation score was `r fmt_num(case1_tnie$Estimate[1])` (`r case1_tnie_ci`), so the mediator pathway itself still points in the same harmful direction
- the estimated proportion mediated was `r fmt_pct(case1_pm$Estimate[1])` (`r case1_pm_ci`), suggesting that inflammation may explain a sizeable share of the total association in this synthetic cohort

Scientifically, that is the pattern we would describe as:

> higher metabolic risk is associated with both a higher inflammation burden
> and a higher incident hazard, and the mediation decomposition is compatible
> with inflammation carrying a meaningful part of that pathway.

The pathway figure makes the same result easier to communicate in a paper or
slide deck.

```{r, fig.width = 8.6, fig.height = 5.8, dev.args = list(bg = "white")}
leo_cox_mediation_plot(
  x = res_case1,
  exposure_label = "Metabolic\nrisk",
  mediator_label = "Inflammation\nscore",
  outcome_label = "Incident\noutcome",
  language = "en",
  palette = "jama"
)
```

## Case 2. The same biology, but a thresholded binary mediator

Sometimes the scientific question is about a thresholded inflammatory state
rather than a graded biomarker. The model then changes from `mreg = "linear"`
to `mreg = "logistic"`.

```{r}
res_case2 <- leo_cox_mediation(
  df = rare_case,
  y_out = c("incident_event", "followup_years"),
  x_exp = "metabolic_risk",
  x_med = "inflammation_group",
  x_cov = c("age", "sex", "smoking"),
  x_exp_a0 = 0,
  x_exp_a1 = 1,
  x_med_cde = 1,
  x_cov_cond = list(age = 60, sex = "Female", smoking = "Never"),
  mreg = "logistic",
  yreg = "auto",
  interaction = TRUE,
  verbose = FALSE
)

case2_tnie <- get_effect_row(res_case2, "tnie")
case2_pm <- get_effect_row(res_case2, "pm")

dplyr::bind_rows(
  case_summary_row(res_case1, "Case 1: rare outcome + continuous mediator"),
  case_summary_row(res_case2, "Case 2: rare outcome + binary mediator")
)
```

The scientific lesson is not that binary mediators are wrong. It is that the
mediator encoding should match the question:

- in Case 1, the indirect pathway was `r fmt_num(case1_tnie$Estimate[1])` on the hazard-ratio scale, with `pm` estimated at `r fmt_pct(case1_pm$Estimate[1])`
- in Case 2, the same broad story became much weaker and less precise (`tnie` = `r fmt_num(case2_tnie$Estimate[1])`, `pm` = `r fmt_pct(case2_pm$Estimate[1])`)

That is a useful teaching contrast. If the biology is genuinely graded, turning
the mediator into a yes/no label can discard pathway information. If the real
question is threshold-based, however, the binary mediator may be the better
scientific choice even if it is statistically less efficient.

## Case 3. The same pathway question, but with a more common outcome

Now keep the same exposure-mediator story and only make the event more common.
This is where `yreg = "auto"` becomes important.

```{r}
res_case3 <- leo_cox_mediation(
  df = common_case,
  y_out = c("incident_event", "followup_years"),
  x_exp = "metabolic_risk",
  x_med = "inflammation_score",
  x_cov = c("age", "sex", "smoking"),
  x_exp_a0 = 0,
  x_exp_a1 = 1,
  x_med_cde = stats::median(common_case$inflammation_score, na.rm = TRUE),
  x_cov_cond = list(age = 60, sex = "Female", smoking = "Never"),
  mreg = "linear",
  yreg = "auto",
  interaction = TRUE,
  verbose = FALSE
)

case3_te <- get_effect_row(res_case3, "te")
case3_tnie <- get_effect_row(res_case3, "tnie")
case3_pm <- get_effect_row(res_case3, "pm")

res_case3$evaluation[, c("event_rate", "mreg", "yreg", "interaction", "x_cov_cond")]
dplyr::filter(res_case3$result, `Effect code` %in% c("te", "tnie", "pm"))
```

Here the event rate was `r fmt_pct(res_case3$evaluation$event_rate[1])`, so
`yreg = "auto"` switched to `r res_case3$evaluation$yreg[1]`.

That changes the interpretation scale:

- `te` = `r fmt_num(case3_te$Estimate[1])` is now a time ratio, not a hazard ratio
- because the time ratio is below 1, the higher-risk group has *shorter* event-free time
- `tnie` = `r fmt_num(case3_tnie$Estimate[1])` means the inflammation pathway is also pointing toward shorter event-free time

This is the same biological story on a different model scale. The key practical
point is that the wording must change with `yreg`:

- `survCox`: talk about higher or lower hazard
- `survAFT_weibull`: talk about shorter or longer event-free time

## Put the three scenarios side by side

```{r}
dplyr::bind_rows(
  case_summary_row(res_case1, "Case 1: rare outcome + continuous mediator"),
  case_summary_row(res_case2, "Case 2: rare outcome + binary mediator"),
  case_summary_row(res_case3, "Case 3: common outcome + continuous mediator")
)
```

This comparison is useful because it separates three different decisions:

1. how to define the mediator scientifically
2. how much information is lost when a graded mediator is dichotomized
3. how the interpretation scale changes when the outcome is no longer rare

## Key practical takeaways

If you only keep six practical rules from Step 3.1, make them these:

1. choose `x_exp_a0` and `x_exp_a1` to match the exposure contrast you would actually write in a Results section
2. choose `x_med_cde` to represent the mediator level at which you want the controlled direct effect to be interpreted; for continuous biomarkers, the median is often a clean first choice
3. choose `x_cov_cond` to represent a real covariate profile, especially when `x_cov` includes factor variables
4. use `mreg = "linear"` when the mediator is scientifically graded, and `mreg = "logistic"` when the mediator question is genuinely threshold-based
5. let `yreg = "auto"` protect you from using `survCox` in obviously non-rare settings, but always read the final `Outcome model` before interpreting the ratio scale
6. write the scientific conclusion from the actual fitted output, not from the existence of a pathway diagram or a single positive point estimate
