---
title: "Survival Analysis Step 3.1: Mediation Case Study"
description: "Move from the default mediation route to the advanced regmedint workflow for survival mediation case studies."
output:
  rmarkdown::html_vignette:
    highlight: default
vignette: >
  %\VignetteIndexEntry{Survival Analysis Step 3.1: Mediation Case Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(leo.ukb))
has_regmedint <- requireNamespace("regmedint", quietly = TRUE)
if (!has_regmedint) {
  knitr::opts_chunk$set(eval = FALSE)
}
fmt_num <- function(x, digits = 3) {
  if (length(x) == 0 || all(is.na(x))) return("NA")
  sprintf(paste0("%.", digits, "f"), x[1])
}
fmt_pct <- function(x, digits = 1) {
  if (length(x) == 0 || all(is.na(x))) return("NA")
  sprintf(paste0("%.", digits, "f%%"), 100 * x[1])
}
effect_row <- function(res, code) {
  if (is.null(res)) return(data.frame())
  dplyr::filter(res$result, `Effect code` == code)
}
effect_value <- function(res, code, column = "Estimate") {
  row <- effect_row(res, code)
  if (!nrow(row) || !column %in% names(row)) return(NA_real_)
  row[[column]][1]
}
effect_text <- function(res, code, column = "95% CI") {
  row <- effect_row(res, code)
  if (!nrow(row) || !column %in% names(row)) return("NA")
  as.character(row[[column]][1])
}
eval_value <- function(res, column) {
  if (is.null(res) || !column %in% names(res$evaluation)) return(NA)
  res$evaluation[[column]][1]
}
fmt_pct_ci <- function(ci_text, digits = 1) {
  ci_num <- suppressWarnings(as.numeric(trimws(strsplit(ci_text, ",", fixed = TRUE)[[1]])))
  if (length(ci_num) != 2 || anyNA(ci_num)) return(ci_text)
  paste(sprintf(paste0("%.", digits, "f%%"), 100 * ci_num), collapse = ", ")
}
case_summary_row <- function(res, scenario) {
  te_code <- if ("te" %in% res$result$`Effect code`) "te" else "ate"
  tnie_code <- if ("tnie" %in% res$result$`Effect code`) "tnie" else "acme"
  pm_code <- "pm"
  te <- effect_row(res, te_code)
  tnie <- effect_row(res, tnie_code)
  pm <- effect_row(res, pm_code)
  data.frame(
    Scenario = scenario,
    `Event rate` = fmt_pct(eval_value(res, "event_rate")),
    `Outcome model` = eval_value(res, "yreg"),
    `Total effect` = sprintf("%s (%s)", te$Estimate[1], te$`95% CI`[1]),
    `Indirect effect` = sprintf("%s (%s)", tnie$Estimate[1], tnie$`95% CI`[1]),
    `Proportion mediated` = sprintf("%s (%s)", fmt_pct(pm$Estimate[1]), fmt_pct_ci(pm$`95% CI`[1])),
    check.names = FALSE
  )
}
draw_mediation_map <- function(language = c("en", "zh")) {
  language <- match.arg(language)
  exposure_label <- if (language == "en") "Exposure (X)" else "暴露 (X)"
  mediator_label <- if (language == "en") "Mediator (M)" else "中介 (M)"
  outcome_label <- if (language == "en") "Outcome (Y)" else "结局 (Y)"
  direct_label <- if (language == "en") "Direct path:\nPNDE / TNDE" else "直接路径：\nPNDE / TNDE"
  indirect_label <- if (language == "en") "Indirect path:\nPNIE / TNIE" else "间接路径：\nPNIE / TNIE"
  total_label <- if (language == "en") "Overall contrast:\nTE" else "总体比较：\nTE"
  cde_label <- if (language == "en") {
    "CDE: direct effect when M is fixed at x_med_cde"
  } else {
    "CDE: direct effect when M is fixed at x_med_cde"
  }

  graphics::plot.new()
  graphics::plot.window(xlim = c(0, 10), ylim = c(0, 10))
  graphics::rect(0.8, 4.4, 3.0, 6.6, col = "#DCEBFF", border = "#4A76A8", lwd = 1.3)
  graphics::rect(3.9, 7.0, 6.1, 9.2, col = "#FDE6D8", border = "#C96E3D", lwd = 1.3)
  graphics::rect(7.0, 4.4, 9.2, 6.6, col = "#E3F4E8", border = "#5B8A72", lwd = 1.3)
  graphics::text(1.9, 5.5, exposure_label, cex = 1)
  graphics::text(5.0, 8.1, mediator_label, cex = 1)
  graphics::text(8.1, 5.5, outcome_label, cex = 1)
  graphics::arrows(3.05, 5.5, 6.95, 5.5, length = 0.09, lwd = 2, col = "#4A76A8")
  graphics::arrows(3.0, 6.3, 3.95, 7.7, length = 0.09, lwd = 2, col = "#C96E3D")
  graphics::arrows(6.05, 7.7, 7.0, 6.3, length = 0.09, lwd = 2, col = "#C96E3D")
  graphics::text(5.0, 4.65, direct_label, cex = 0.95)
  graphics::text(5.0, 6.95, indirect_label, cex = 0.95)
  graphics::text(5.0, 2.55, total_label, cex = 0.95)
  graphics::segments(1.9, 3.3, 8.1, 3.3, lwd = 1.5, col = "#6B6B6B")
  graphics::segments(1.9, 3.3, 1.9, 3.0, lwd = 1.5, col = "#6B6B6B")
  graphics::segments(8.1, 3.3, 8.1, 3.0, lwd = 1.5, col = "#6B6B6B")
  graphics::text(5.0, 1.55, cde_label, cex = 0.92)
  graphics::box()
}
res_case1 <- NULL
res_case2 <- NULL
res_case3 <- NULL
```

## Why this advanced article exists

Step 3.0 is the default path: `method = "mediation"`, average-effect outputs,
and a careful interpretation on the time scale. Step 3.1 is for the moment
when that default route is no longer the question you want to answer.

Move to `method = "regmedint"` when you specifically need one or more of the
following:

1. the `TE / TNIE / PNDE / PM` family rather than `ACME / ADE / ATE`
2. a controlled direct effect evaluated at `x_med_cde`
3. explicit `survCox` versus `survAFT_*` handling
4. a decomposition that is written around a declared exposure contrast and a
   declared covariate profile

In other words, this is the article for the analysis that is being written
closer to a methods section, not just to a first-pass teaching example.

```{r, echo = FALSE, results = "asis"}
if (!has_regmedint) {
  cat(
    "> Note: package `regmedint` is not installed in the current build environment, ",
    "so the code is shown but not executed. Install `regmedint` locally to run ",
    "the full case study.\n\n",
    sep = ""
  )
}
```

## Keep one scientific story and vary only the analytical pressure

The scientific question stays the same across all scenarios:

> is the association between higher metabolic risk and an incident outcome
> partly transmitted through inflammation?

What changes across the cases is not the story, but the analytical pressure we
put on that story:

- continuous versus binary mediator
- rare versus common outcome
- Cox versus AFT interpretation

```{r}
set.seed(20260330)

make_mediation_case <- function(n = 1200, base_log_hazard = -5.8, censor_rate = 0.06) {
  age <- rnorm(n, 60, 8)
  sex <- factor(sample(c("Female", "Male"), n, replace = TRUE))
  smoking <- factor(sample(c("Never", "Ever"), n, replace = TRUE, prob = c(0.6, 0.4)))
  sex_num <- as.integer(sex == "Male")
  smoking_num <- as.integer(smoking == "Ever")
  metabolic_risk_num <- rbinom(
    n, 1,
    plogis(-0.3 + 0.03 * (age - 60) + 0.30 * sex_num + 0.35 * smoking_num)
  )
  inflammation_score <- rnorm(
    n,
    mean = 0.85 * metabolic_risk_num + 0.04 * (age - 60) + 0.35 * smoking_num,
    sd = 0.9
  )
  hazard_rate <- exp(
    base_log_hazard +
      0.55 * metabolic_risk_num +
      0.24 * inflammation_score +
      0.02 * (age - 60) +
      0.20 * smoking_num
  )
  event_time <- rexp(n, rate = hazard_rate)
  censor_time <- rexp(n, rate = censor_rate)

  data.frame(
    incident_event = as.integer(event_time <= censor_time),
    followup_years = pmin(event_time, censor_time),
    metabolic_risk = factor(
      metabolic_risk_num,
      levels = c(0, 1),
      labels = c("Lower risk", "Higher risk")
    ),
    inflammation_score = inflammation_score,
    inflammation_group = factor(
      as.integer(inflammation_score > 0.5),
      levels = c(0, 1),
      labels = c("Lower inflammation", "Higher inflammation")
    ),
    age = age,
    sex = sex,
    smoking = smoking
  )
}

rare_case <- make_mediation_case(base_log_hazard = -5.8, censor_rate = 0.06)
common_case <- make_mediation_case(base_log_hazard = -4.6, censor_rate = 0.06)

scenario_overview <- data.frame(
  Scenario = c("Rare-outcome cohort", "Common-outcome cohort"),
  Participants = c(nrow(rare_case), nrow(common_case)),
  Events = c(sum(rare_case$incident_event), sum(common_case$incident_event)),
  `Event rate` = c(
    sprintf("%.1f%%", 100 * mean(rare_case$incident_event)),
    sprintf("%.1f%%", 100 * mean(common_case$incident_event))
  )
)

knitr::kable(
  scenario_overview,
  caption = "The two core cohorts used in the Step 3.1 case study"
)
```

## Keep the pathway sketch in mind before looking at the numbers

```{r, echo = FALSE}
draw_mediation_map("en")
```

The point of this figure is not artistic. It keeps the vocabulary stable while
the scenarios change:

- `TE` is the total contrast
- `TNIE` is the indirect pathway we usually look at first
- `PNDE` is the direct pathway that bypasses the mediator
- `CDE` becomes useful only after you explicitly choose `x_med_cde`

That explicitness is the main difference from Step 3.0.

That is why a good teacher would usually make students say the abbreviations
out loud before talking about the numbers. If the vocabulary is still shaky,
the interpretation of `TNIE` or `PM` is usually shaky too.

## Case 1. Rare outcome, continuous mediator, explicit evaluation profile

We start with the most common advanced use case: a graded biomarker mediator,
a rare outcome, and a declared covariate profile.

```{r}
res_case1 <- leo_cox_mediation(
  df = rare_case,
  y_out = c("incident_event", "followup_years"),
  x_exp = "metabolic_risk",
  x_med = "inflammation_score",
  x_cov = c("age", "sex", "smoking"),
  method = "regmedint",
  x_exp_a0 = 0,
  x_exp_a1 = 1,
  x_med_cde = stats::median(rare_case$inflammation_score, na.rm = TRUE),
  x_cov_cond = list(age = 60, sex = "Female", smoking = "Never"),
  mreg = "linear",
  yreg = "auto",
  verbose = FALSE
)

res_case1$evaluation[, c(
  "exposure_contrast", "x_exp_a0", "x_exp_a1",
  "x_med_cde", "x_cov_cond", "mreg", "yreg", "interaction"
)]
```

One useful subtlety appears already: because `interaction = NULL` resolves to
`TRUE` for `method = "regmedint"`, the advanced route is more interaction-ready
by default than the Step 3.0 route. If your scientific question does not want
that, say so explicitly with `interaction = FALSE`.

```{r}
res_case1$result[res_case1$result$`Effect code` %in% c("cde", "pnde", "tnie", "te", "pm"), ]
```

```{r}
dplyr::select(
  res_case1$result[res_case1$result$`Effect code` %in% c("cde", "pnde", "tnie", "te", "pm"), ],
  `Effect code`, Scale, `Exposure contrast`, `Covariate profile`,
  `Mediator reference`, Estimate, `95% CI`
)
```

This is the teaching readout:

- the total effect is `r fmt_num(effect_value(res_case1, "te"))`
  (`r effect_text(res_case1, "te")`)
- the indirect effect `TNIE` is `r fmt_num(effect_value(res_case1, "tnie"))`
  (`r effect_text(res_case1, "tnie")`)
- the proportion mediated is `r fmt_pct(effect_value(res_case1, "pm"))`
  (`r fmt_pct_ci(effect_text(res_case1, "pm"))`)

Because `yreg = survCox` here, these non-PM rows are read on the hazard-ratio
scale. In plain language, this case says that the higher-risk exposure contrast
is associated with higher hazard overall, and a meaningful part of that
contrast is compatible with flowing through the inflammation pathway.

The disciplined reading order is therefore not "estimate first." It is:
contrast first, profile second, scale third, estimate fourth. That is the
difference between merely printing the advanced route and actually teaching it.

Put differently, if a student begins with "the TNIE is ..." but cannot yet say
who is being compared, under which profile, and on which scale, the answer is
not methodologically complete. The value of the advanced route is not that it
has more symbols; it is that the comparison is specified more honestly.

The scientific boundary is just as important as the estimate. This Cox-based
decomposition is most comfortable when the event remains relatively rare; in
this scenario the event rate is `r fmt_pct(eval_value(res_case1, "event_rate"))`,
which is why `yreg = "auto"` stays on `survCox`.

## Case 2. Same rare outcome, but now the mediator is binary

Many biomarker papers eventually dichotomize a mediator for clinical
communication. Once you do that, the main new obligation is to define
`x_med_cde` in the observed mediator coding.

```{r}
res_case2 <- leo_cox_mediation(
  df = rare_case,
  y_out = c("incident_event", "followup_years"),
  x_exp = "metabolic_risk",
  x_med = "inflammation_group",
  x_cov = c("age", "sex", "smoking"),
  method = "regmedint",
  x_exp_a0 = 0,
  x_exp_a1 = 1,
  x_med_cde = "Higher inflammation",
  x_cov_cond = list(age = 60, sex = "Female", smoking = "Never"),
  mreg = "logistic",
  yreg = "auto",
  verbose = FALSE
)

res_case2$evaluation[, c(
  "exposure_contrast", "x_med_cde", "mediator_reference",
  "mreg", "yreg", "interaction"
)]
```

```{r}
res_case2$result[res_case2$result$`Effect code` %in% c("cde", "pnde", "tnie", "te", "pm"), ]
```

```{r}
dplyr::select(
  res_case2$result[res_case2$result$`Effect code` %in% c("cde", "pnde", "tnie", "te", "pm"), ],
  `Effect code`, `Mediator reference`, Estimate, `95% CI`
)
```

Here the `CDE` row becomes more tangible than in the continuous-mediator case,
because it now answers a concrete question such as:

> if two otherwise similar participants were both fixed at the "Higher
> inflammation" mediator level, how different would their direct exposure
> contrast look?

That is why `x_med_cde` belongs in the advanced route. It is not a nuisance
parameter. It is part of the scientific question.

For the same reason, `x_med_cde` should never be written vaguely. The mediator
level at which it is fixed is part of the biological or clinical scenario you
are claiming to describe.

## Case 3. A commoner outcome pushes the same question onto the AFT path

Now we keep the biological story, return to the continuous mediator, but make
the event common enough that `yreg = "auto"` no longer stays on Cox.

```{r}
res_case3 <- leo_cox_mediation(
  df = common_case,
  y_out = c("incident_event", "followup_years"),
  x_exp = "metabolic_risk",
  x_med = "inflammation_score",
  x_cov = c("age", "sex", "smoking"),
  method = "regmedint",
  x_exp_a0 = 0,
  x_exp_a1 = 1,
  x_med_cde = stats::median(common_case$inflammation_score, na.rm = TRUE),
  x_cov_cond = list(age = 60, sex = "Female", smoking = "Never"),
  mreg = "linear",
  yreg = "auto",
  verbose = FALSE
)

res_case3$evaluation[, c("event_rate", "yreg", "outcome_distribution", "interaction")]
```

```{r}
res_case3$result[res_case3$result$`Effect code` %in% c("cde", "pnde", "tnie", "te", "pm"), ]
```

```{r}
dplyr::select(
  res_case3$result[res_case3$result$`Effect code` %in% c("cde", "pnde", "tnie", "te", "pm"), ],
  `Effect code`, Scale, `Outcome model`, Estimate, `95% CI`
)
```

This is the place where many readers get turned around. Once `yreg = auto`
switches to `survAFT_weibull`, the non-PM rows are now read as **time ratios**:

- `TR > 1` is compatible with longer event-free time
- `TR < 1` is compatible with shorter event-free time

So the numerical summary has not become "wrong" or "inconsistent." It is
answering the same scientific question on a different outcome-model scale.

## Put the three cases side by side

```{r}
summary_table <- dplyr::bind_rows(
  case_summary_row(res_case1, "Rare outcome + continuous mediator"),
  case_summary_row(res_case2, "Rare outcome + binary mediator"),
  case_summary_row(res_case3, "Common outcome + continuous mediator")
)

knitr::kable(
  summary_table,
  caption = "How the same scientific question changes across three advanced mediation settings"
)
```

Lined up this way, the methodological lesson becomes much cleaner:

- changing the mediator encoding changes the mediator model and the meaning of
  `x_med_cde`
- changing the event frequency can change the outcome-model family
- neither change is cosmetic; both alter how the effect sizes should be read

For many students, that is the real turning point of the advanced tutorial.
The lesson is not "there are more abbreviations now." It is that almost every
column in the advanced result table is helping determine whether the final
sentence will be scientifically correct.

## Draw the advanced pathway figure

```{r}
leo_cox_mediation_plot(
  x = res_case1,
  exposure_label = "Metabolic\nrisk",
  mediator_label = "Inflammation\nscore",
  outcome_label = "Incident\noutcome",
  language = "en",
  palette = "lancet"
)
```

For the advanced route, the plot is most useful after you have already read the
evaluation table. Otherwise the labels look tidy but the scientific comparison
remains under-specified.

## What Step 3.1 should leave you with

The advanced route is not "better" because it is longer. It is better only
when the question actually requires it.

Use `method = "regmedint"` when you need to say all of the following with a
straight face:

1. which exposure contrast is being decomposed
2. which mediator value defines the controlled direct effect
3. which covariate profile defines the evaluation point
4. whether the outcome-model scale is Cox HR or AFT time ratio

If you do not need all four, Step 3.0 is often the more disciplined analysis.
If you do need them, Step 3.1 is where the decomposition becomes explicit
enough for a serious methods paragraph.
