---
title: "Survival Analysis Step 2: Subgroup and Interaction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survival Analysis Step 2: Subgroup and Interaction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(leo.ukb))
```

# What this step is for

This tutorial is the second survival-analysis step in `leo.ukb`. After fitting
the main Cox model, a common next question is:

> does the exposure effect differ across groups?

This is where subgroup analysis and interaction analysis begin.

# First separate two questions

Subgroup analysis and interaction analysis are related, but they do not answer
exactly the same question.

- `leo_cox_subgroup()` asks: what is the hazard ratio within each subgroup?
- `leo_cox_interaction()` asks: are those subgroup-specific effects genuinely
  different from each other on the multiplicative hazard-ratio scale?

This distinction matters in papers. One subgroup can be "statistically
significant" and another can be "not significant" without there being strong
evidence that the two subgroup effects differ.

# Build a compact teaching data set

We continue with the `survival::lung` example.

```{r}
lung_df <- dplyr::transmute(
  survival::lung,
  outcome = as.integer(status == 2),
  outcome_censor = time / 365.25,
  age = age,
  sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
  ecog_group = factor(
    ph.ecog,
    levels = 0:3,
    labels = c("ECOG0", "ECOG1", "ECOG2", "ECOG3")
  )
)

cox_models <- list(
  "Crude" = NULL,
  "Model A" = c("ecog_group")
)
```

# Interaction first: the formal heterogeneity test

When the scientific question is "does the effect differ across groups?", the
formal answer comes from an interaction model, not from visually comparing
subgroup-specific `P values`.

```{r}
res_inter <- leo_cox_interaction(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_inter = "sex",
  x_cov = cox_models,
  verbose = FALSE
)

res_inter$result
```

Read this table as the formal interaction summary:

- `P for interaction` is the key heterogeneity test
- in the current Cox workflow, this is a multiplicative interaction test on the
  hazard-ratio scale
- `Interaction class` tells you whether the modifier was treated as binary,
  multi-level categorical, or continuous

For a continuous modifier, `leo_cox_interaction()` can still test interaction,
but `leo_cox_subgroup()` is not the right follow-up unless you first define a
meaningful categorical grouping.

# Subgroup analysis next: show the stratum-specific HRs

After the formal interaction test, it is often useful to show the hazard ratio
within each subgroup itself.

```{r}
res_sub <- leo_cox_subgroup(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_subgroup = c("sex", "ecog_group"),
  x_cov = list("Crude" = NULL),
  verbose = FALSE
)

utils::head(res_sub$result, 8)
```

The subgroup table is designed to be easy to read in a manuscript:

- one row per subgroup level and model
- the subgroup-specific `HR`, `95% CI`, and `P value`
- the matching `P for interaction`

# How to interpret the two outputs together

The cleanest reporting logic is:

1. use `leo_cox_interaction()` to decide whether there is evidence that the
   exposure effect differs by subgroup
2. use `leo_cox_subgroup()` to describe the pattern of subgroup-specific hazard
   ratios

A practical interpretation rule is:

- `P for interaction` is the primary evidence for heterogeneity
- subgroup-specific `HR` values describe the direction and magnitude within
  each group

This is why subgroup analysis should not be reduced to "one subgroup was
significant and another was not".

# How to write the finding in a paper

In a manuscript, the safest reporting order is:

1. state whether the interaction test supported heterogeneity
2. then describe the subgroup-specific hazard ratios

For example:

> The association between age and the event differed by sex
> (`P for interaction = ...`). In subgroup analyses, the hazard ratio was
> higher in ... than in ...

This wording keeps the formal test and the descriptive pattern in the right
order.

# When should subgroup variables be categorical?

`leo_cox_subgroup()` is designed for categorical subgroup variables.

- binary subgroup variables are natural here, such as `sex`
- multi-level subgroup variables are also fine, such as `ecog_group`
- continuous modifiers belong first in `leo_cox_interaction()`

If a modifier is continuous, you have two options:

1. keep it continuous and do a formal interaction test only
2. define a clinically meaningful grouping first, then use subgroup analysis

The choice should follow the scientific question, not convenience.

# Optional heterogeneity P values

By default, `leo_cox_subgroup()` appends `P for interaction`. This is the main
heterogeneity signal for most Cox subgroup tables in cohort papers.

`P for heterogeneity` is available as an optional supplement:

```{r}
res_sub_hetero <- leo_cox_subgroup(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_subgroup = "sex",
  x_cov = list("Crude" = NULL),
  add_heterogeneity = TRUE,
  verbose = FALSE
)

res_sub_hetero$result
```

In most cohort papers, it is better to treat `P for heterogeneity` as
supporting information rather than the primary decision point.

# What to take away from step 2

This step should leave you with three practical rules:

1. subgroup analysis and interaction analysis are connected, but not identical
2. `P for interaction` is the formal test of effect modification
3. subgroup-specific hazard ratios are for description, not for replacing the
   interaction test

With that logic in place, your survival-analysis reporting becomes much easier
to defend scientifically.

Step 3 then asks a new question:

> could part of the exposure-outcome association operate through a mediator?
