---
title: "Survival Analysis Step 2: Subgroup and Interaction"
description: "Learn how to choose, run, and interpret subgroup, multiplicative interaction, and additive interaction analyses for survival outcomes."
output:
  rmarkdown::html_vignette:
    highlight: default
vignette: >
  %\VignetteIndexEntry{Survival Analysis Step 2: Subgroup and Interaction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(leo.ukb))
```

# What this step is for

This tutorial is the second survival-analysis step in `leo.ukb`. After the
main Cox model, the next practical question is usually:

> does the exposure-outcome association differ across groups, and if so, on
> what scale?

That question can lead to three related but different outputs:

1. a formal multiplicative interaction test
2. a subgroup table that describes the pattern of hazard ratios
3. an additive-interaction summary for binary-binary contrasts

The key is not to mix them up. Step 2 is where we learn how to choose the
right one and how to read the results without over-interpreting them.

# Start from the data and analytic roles

We keep using `survival::lung`, but this time we build the data frame around
teaching roles rather than only model fitting.

```{r}
lung_df <- dplyr::transmute(
  survival::lung,
  outcome = as.integer(status == 2),
  outcome_censor = time / 365.25,
  age = age,
  age_high = factor(
    ifelse(age >= 65, "Older", "Younger"),
    levels = c("Younger", "Older")
  ),
  sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
  ecog_group3 = factor(
    dplyr::case_when(
      ph.ecog == 0 ~ "ECOG0",
      ph.ecog == 1 ~ "ECOG1",
      ph.ecog >= 2 ~ "ECOG2plus",
      TRUE ~ NA_character_
    ),
    levels = c("ECOG0", "ECOG1", "ECOG2plus")
  ),
  ecog_binary = factor(
    ifelse(ph.ecog >= 2, "Poor", "Good"),
    levels = c("Good", "Poor")
  )
)
```

Before fitting anything, look at the structure the way you would in a real
paper:

```{r}
dplyr::glimpse(lung_df)
```

Table: Analytic roles in the Step 2 teaching data frame

| Variable | Type | Teaching role |
| --- | --- | --- |
| `outcome` | Binary event indicator | Outcome event |
| `outcome_censor` | Follow-up time in years | Outcome time |
| `age` | Continuous exposure | Main exposure in multiplicative examples |
| `sex` | Binary modifier | Binary modifier for multiplicative interaction |
| `ecog_group3` | 3-level ECOG performance-status grouping | Multi-level modifier for subgroup display |
| `age_high` | Binary age grouping | Binary age variable for additive interaction |
| `ecog_binary` | Binary ECOG performance-status grouping | Binary ECOG modifier for additive interaction |

It is also worth checking the subgroup and joint-cell sizes before talking
about effect modification.

```{r}
knitr::kable(
  dplyr::count(lung_df, ecog_group3, outcome),
  caption = "Outcome counts across the multi-level subgroup"
)
```

```{r}
knitr::kable(
  dplyr::count(lung_df, age_high, ecog_binary),
  caption = "Four joint cells for the additive-interaction example"
)
```

The second table is especially important because additive interaction needs a
binary-by-binary setup with all four cells present.

# Choose the scale before fitting

The easiest way to avoid confusion is to decide what question you are asking
before you pick the function.

```{r}
decision_table <- data.frame(
  Goal = c(
    "Formal multiplicative interaction",
    "Subgroup display",
    "Additive interaction"
  ),
  Exposure_types = c(
    "Continuous, binary, or multi-level categorical",
    "Continuous, binary, or categorical",
    "Binary only"
  ),
  Modifier_types = c(
    "Continuous, binary, or multi-level categorical",
    "Binary or multi-level categorical",
    "Binary only"
  ),
  Main_function = c(
    "leo_cox_interaction()",
    "leo_cox_subgroup()",
    "leo_cox_add_interaction()"
  ),
  Main_question = c(
    "Does the association differ on the hazard-ratio scale?",
    "What does the HR look like within each stratum?",
    "Does the joint effect exceed what we expect under additivity?"
  )
)

knitr::kable(
  decision_table,
  caption = "Which Step 2 tool matches which scientific question?"
)
```

This table is the practical rule of thumb:

- multiplicative interaction is the most general option
- subgroup tables need a categorical modifier because they need explicit strata
- additive interaction is currently the most restrictive because `RERI`, `AP`,
  and `S` require a binary-by-binary contrast

If a modifier is continuous, `leo_cox_interaction()` can still test it
formally, but `leo_cox_subgroup()` is only appropriate after a clinically
meaningful grouping has been defined.

# Multiplicative interaction is the general formal test

In a Cox model, multiplicative interaction asks whether the exposure effect
differs across levels of a modifier on the hazard-ratio scale.

## Example 1: continuous exposure by binary modifier

Here we ask whether the age-outcome association differs by sex.

```{r}
res_inter_sex <- leo_cox_interaction(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_inter = "sex",
  x_cov = list(
    "Crude" = NULL,
    "Model A" = c("ecog_group3")
  ),
  verbose = FALSE
)

res_inter_sex$result
```

How to read this result:

- `Exposure class = Continuous` means `age` is handled as a continuous
  exposure
- `Interaction class = Binary` means `sex` is treated as a binary modifier
- `P for interaction` is the formal heterogeneity test

In this teaching example, the adjusted `P for interaction` is `0.18`, so there
is no clear evidence that the age-outcome association differs by sex on the
multiplicative hazard-ratio scale.

That is already a useful result. A non-significant interaction does not mean
"nothing happened"; it means the data do not strongly support effect
modification on this scale.

## Example 2: continuous exposure by multi-level categorical modifier

The same function also works for a multi-level subgroup variable.

```{r}
res_inter_ecog <- leo_cox_interaction(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_inter = "ecog_group3",
  x_cov = list("Crude" = NULL),
  verbose = FALSE
)

res_inter_ecog$result
```

Here the modifier is a 3-level categorical variable, and the interaction test
returns `P for interaction = 0.037`.

That means the age-outcome association appears to differ across ECOG groups on
the multiplicative hazard-ratio scale. Because the modifier is categorical, a
subgroup table is now a natural follow-up.

# Subgroup analysis describes the pattern after the formal test

`leo_cox_subgroup()` is not a replacement for the interaction test. It is the
display layer that helps you describe what the pattern looks like once a
categorical modifier has been chosen.

```{r}
res_sub_ecog <- leo_cox_subgroup(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_subgroup = "ecog_group3",
  x_cov = list("Crude" = NULL),
  verbose = FALSE
)

res_sub_ecog$result
```

This subgroup table tells a more descriptive story:

- in `ECOG0`, the HR per one-year increase in age is `1.059`
- in `ECOG1`, the HR is close to null at `1.004`
- in `ECOG2plus`, the HR is also close to null at `0.994`
- the matching `P for interaction = 0.037` reminds us that the subgroup
  pattern should be interpreted through the formal interaction test

In manuscript language, the safest interpretation is:

> There was evidence that the age-outcome association differed by ECOG
> performance status on the multiplicative scale. The age-associated hazard
> increase appeared stronger in `ECOG0` than in `ECOG1` or `ECOG2plus`.

This is much better than writing "one subgroup was significant and the others
were not", which is not a proper interaction argument.

# Additive interaction answers a different question

Multiplicative and additive interaction are not the same. Additive interaction
asks whether the joint effect of two binary factors is more or less than what
we would expect from adding their separate effects.

That is why the current teaching workflow in `leo.ukb` is binary-by-binary.

Here we use:

- `ecog_binary`: poorer versus better ECOG performance status
- `age_high`: older versus younger age

The binary age grouping is only for teaching additive interaction. In a real
analysis, the cutpoint should be clinically justified in advance rather than
chosen after looking at the results.

```{r}
res_add <- leo_cox_add_interaction(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "ecog_binary",
  x_inter = "age_high",
  x_cov = list("Crude" = NULL),
  verbose = FALSE
)

res_add$result
```

This output should be read in layers.

First, the setup columns:

- `Reference group` tells you which joint category is treated as the baseline
- `Recode applied` tells you whether the function had to recode the binary
  variables internally so the reference structure was interpretable

Then the scale-specific measures:

- `Multiplicative interaction`: interaction on the hazard-ratio scale
- `RERI`: relative excess risk due to interaction
- `AP`: attributable proportion due to interaction
- `S`: synergy index

The null values differ by measure:

- multiplicative interaction: `1`
- `RERI`: `0`
- `AP`: `0`
- `S`: `1`

In this teaching example:

- the multiplicative interaction estimate is `0.454` with `P for interaction =
  0.029`
- `RERI = -1.475` and `AP = -0.739` both point toward negative interaction on
  the additive scale
- but the confidence intervals for `RERI`, `AP`, and `S` all include their
  null values

The first bullet is easy to misread unless we spell out what the number is.
`Multiplicative interaction` is not the joint-group hazard ratio itself. It is:

`HR11 / (HR10 × HR01)`

So a value below `1` does not mean the joint group is automatically protective
or that the joint-group hazard ratio must be below `1`. It means:

> the observed joint effect is smaller than what we would expect if the two
> separate hazard ratios combined by simple multiplication.

For a concrete example:

- if `ecog_binary` alone had `HR = 2`
- and `age_high` alone had `HR = 3`
- then under no multiplicative interaction the joint-group HR would be expected
  to be `2 × 3 = 6`
- if the multiplicative interaction term were `0.454`, the joint-group HR would
  instead be about `2 × 3 × 0.454 = 2.724`

That joint-group HR is still above `1`; it is just weaker than the pure
multiplicative expectation. In other words, this points to negative
interaction, or antagonism, on the multiplicative scale.

So the most careful interpretation is:

> The binary contrast showed evidence of interaction on the multiplicative
> scale, but additive-interaction estimates were imprecise and did not provide
> clear evidence on the additive scale.

This is a good teaching result because it shows that the answer can depend on
the scale. A significant multiplicative interaction does not guarantee a clear
additive interaction, and vice versa.

# How to interpret common result patterns

The practical reading rules are:

1. **Multiplicative interaction supported, additive interaction not supported**

   State that effect modification is supported on the hazard-ratio scale, while
   additive-interaction evidence remains uncertain.

2. **Additive interaction supported, multiplicative interaction not supported**

   State that the joint effect exceeds additivity, even though the hazard-ratio
   scale does not show clear multiplicative deviation.

3. **Both supported**

   State explicitly that the association differs on both scales.

4. **Neither supported**

   State that there is no clear evidence of effect modification in the current
   data.

And one rule should always stay in place:

> "one subgroup was significant and another was not" is not itself evidence of
> interaction.

# How to write Step 2 results in a paper

Here are manuscript-style templates that match the outputs above.

For a null multiplicative interaction:

> There was no clear evidence that the association between age and the event
> differed by sex after adjustment (`P for interaction = 0.18`).

For a supported multiplicative interaction followed by a subgroup table:

> The association between age and the event differed by ECOG performance status
> (`P for interaction = 0.037`). In subgroup analyses, the hazard ratio per
> one-year increase in age was `1.059` (`95% CI 1.013, 1.107`) in `ECOG0`, but
> was close to null in `ECOG1` and `ECOG2plus`.

For additive interaction:

> For the binary contrast between poor performance status and older age,
> the multiplicative interaction term was `0.454` (`95% CI 0.226, 0.911`),
> indicating that the joint effect was smaller than expected under pure
> multiplication of the two separate hazard ratios; this does not itself mean
> that the joint-group HR was below `1`. Additive-interaction estimates remained
> imprecise
> (`RERI -1.475`, `95% CI -3.199, 0.249`).

This wording keeps the scale explicit and avoids over-claiming.

# What to take away from Step 2

Step 2 should leave you with five practical rules:

1. multiplicative interaction is the most general formal test in the Cox
   workflow
2. subgroup tables are descriptive and work best when the modifier is
   categorical
3. additive interaction currently belongs to binary-by-binary contrasts
4. the same data can look different on multiplicative and additive scales
5. interpretation should follow the scale actually tested, not only the raw
   subgroup `P values`

Step 3 then asks a different question:

> could part of the exposure-outcome association operate through a mediator?
