---
title: "Survival Analysis Step 1: Main Cox Analysis"
description: "Learn how to prepare, fit, and interpret the main Cox model for an incident survival outcome."
output:
  rmarkdown::html_vignette:
    highlight: default
vignette: >
  %\VignetteIndexEntry{Survival Analysis Step 1: Main Cox Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(leo.ukb))
```

# What this step is for

This tutorial is the first survival-analysis step in `leo.ukb`. Before we talk
about subgroup differences, interaction, or mediation, we need a clean answer
to the first question:

> is there a main exposure-outcome association in the cohort?

That is the role of the main Cox model.

In practice, a strong survival-analysis workflow is easier to defend when it
is split into stages:

1. establish the main exposure-outcome association
2. only then ask whether that association differs across groups
3. only then ask whether part of that association may operate through a
   mediator

Step 1 is therefore the foundation for everything that comes later.

# Start from the cohort question and analytic roles

We use `survival::lung` as a compact teaching example, but we structure the
data frame the same way we would for a real cohort analysis.

```{r}
lung_df <- dplyr::transmute(
  survival::lung,
  outcome = as.integer(status == 2),
  outcome_censor = time / 365.25,
  age = age,
  sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
  ecog_group = factor(
    ph.ecog,
    levels = 0:3,
    labels = c("ECOG0", "ECOG1", "ECOG2", "ECOG3")
  )
)
```

Now check the analytic roles explicitly instead of jumping straight into model
fitting.

Table: Analytic roles in the Step 1 teaching data frame

| Variable | Type | Step 1 role |
| --- | --- | --- |
| `outcome` | Binary event indicator | Incident outcome event |
| `outcome_censor` | Follow-up time in years | Incident outcome follow-up time |
| `age` | Continuous exposure | Main exposure in the first example |
| `sex` | Binary covariate | Basic demographic adjustment |
| `ecog_group` | 4-level ECOG performance-status grouping | ECOG performance-status adjustment |

# Inspect the analysis-ready data before fitting

Main Cox analysis is easier to trust when the cohort structure, event counts,
and follow-up burden have already been checked.

```{r}
dplyr::glimpse(lung_df)
```

```{r}
knitr::kable(utils::head(lung_df, 6))
```

```{r}
event_summary <- data.frame(
  Metric = c("Participants", "Events", "Event rate"),
  Value = c(
    nrow(lung_df),
    sum(lung_df$outcome),
    sprintf("%.1f%%", 100 * mean(lung_df$outcome))
  )
)

knitr::kable(
  event_summary,
  caption = "Overall event burden"
)
```

```{r}
knitr::kable(
  dplyr::count(lung_df, sex, outcome),
  caption = "Event counts by sex"
)
```

```{r}
followup_summary <- summary(lung_df$outcome_censor)
knitr::kable(
  data.frame(
    Statistic = names(followup_summary),
    Value = as.numeric(followup_summary)
  ),
  caption = "Follow-up summary in years"
)
```

These quick checks already tell us whether the event rate, subgroup size, and
follow-up scale look plausible enough for a first-pass Cox analysis.

# Choose the exposure scale before fitting

Before fitting anything, decide whether your exposure is being treated as
continuous or categorical. This decision changes what the hazard ratio means.

```{r}
scale_table <- data.frame(
  Exposure_setup = c("Continuous exposure", "Categorical exposure"),
  Typical_question = c(
    "What is the change in hazard per one-unit increase?",
    "How does each category compare with the reference level?"
  ),
  Example_in_this_step = c("age", "ecog_group"),
  Main_setting = c("Default", "x_exp_type = 'categorical'")
)

knitr::kable(
  scale_table,
  caption = "Choose the exposure scale before fitting the main Cox model"
)
```

This is the practical rule:

- keep the exposure continuous if the scientific question is per-unit change
- treat it as categorical if the scientific question is level-versus-reference
  comparison

# Define the adjustment hierarchy before fitting

In cohort studies, model labels should reflect the scientific adjustment
hierarchy rather than coding convenience alone. `leo_cox()` accepts a named
list, and those names are carried directly into the final result table.

```{r}
cox_models <- list(
  "Crude" = NULL,
  "Model A" = c("sex"),
  "Model B" = c("sex", "ecog_group")
)

cox_models
```

This habit makes the output read much more like a manuscript table from the
beginning.

# Main Cox example 1: continuous exposure

We first fit the age-outcome association.

```{r}
res_cox <- leo_cox(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_cov = cox_models,
  verbose = FALSE
)

res_cox$result
```

# How to read the main Cox result

This output is already paper-oriented:

- `Case N`, `Control N`, and `Person-years` tell you how much information went
  into the fitted model
- `Class` tells you whether the exposure was treated as continuous or
  categorical
- each model block gives `HR`, `95% CI`, and `P value`

For a continuous exposure, the hazard ratio is interpreted per one-unit
increase in the exposure. Here that means per one-year increase in age.

```{r}
age_tidy <- leo_cox_format(res_cox, style = "tidy")
age_model_b <- dplyr::filter(age_tidy, Model == "Model B")
age_model_b
```

In this teaching example, the Model B estimate is `HR = 1.011`, meaning that
after adjustment for sex and ECOG status, the estimated hazard is about
`r sprintf("%.1f", 100 * (age_model_b$HR - 1))`% higher per one-year increase
in age. The confidence interval, however, includes 1, so the adjusted evidence
for a main age-outcome association is limited in this small example.

That is exactly how a main Cox result should be interpreted:

- first the direction and size of the point estimate
- then the uncertainty
- then the scientific caution

# Use tidy output when you need programmable rows

The wide table is convenient for direct reading. The tidy table is better for
row-binding, plotting, or manuscript automation.

```{r}
dplyr::select(
  age_tidy,
  Model, Exposure, Outcome, HR, `95% CI`, `P value`
)
```

A useful rule is:

- use `$result` for manuscript-style reading
- use `leo_cox_format(..., style = "tidy")` for downstream programming

# Main Cox example 2: categorical exposure

The same workflow also handles categorical exposures.

```{r}
res_cox_cat <- leo_cox(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "ecog_group",
  x_cov = list("Crude" = NULL, "Model A" = c("sex")),
  x_exp_type = "categorical",
  verbose = FALSE
)

res_cox_cat$result
```

This table is read differently from the age model:

- `ECOG0` is the reference level
- each other row compares one ECOG level with `ECOG0`
- the interpretation is therefore level-versus-reference rather than per-unit
  increase

```{r}
ecog_tidy <- leo_cox_format(res_cox_cat, style = "tidy")
dplyr::select(
  ecog_tidy,
  Model, Exposure, Level, HR, `95% CI`, `P value`
)
```

In this example, the adjusted Model A hazard ratios suggest that worse
performance status is associated with a higher incident hazard:

- `ECOG1` versus `ECOG0`: `HR = 1.519`
- `ECOG2` versus `ECOG0`: `HR = 2.579`
- `ECOG3` versus `ECOG0`: `HR = 7.756`

The last row also reminds us why raw output must be read critically: `ECOG3`
has only one event in this small teaching data set, so its confidence interval
is extremely wide. Sparse cells are a scientific warning, not just a cosmetic
table issue.

# How to interpret common Step 1 result patterns

For the main Cox model, the usual interpretation logic is:

1. if the HR is clearly away from 1 and the confidence interval does not cross
   1, the main association is both directionally clear and statistically
   supported
2. if the HR is away from 1 but the confidence interval crosses 1, the data
   suggest a possible association, but the estimate is still imprecise
3. if the point estimate is close to 1, the evidence for a main association is
   weak regardless of the exact P value
4. for categorical exposures, always read both the effect size and the cell
   sizes; sparse strata can generate unstable large HRs

Step 1 is therefore not only about “running Cox”. It is about learning how to
separate signal, uncertainty, and sparse-data artifacts.

# How to write Step 1 results in a paper

For a continuous exposure, a concise Results-style sentence could be:

> In the fully adjusted model, each one-year increase in age was associated
> with an estimated 1.1% higher incident hazard (`HR = 1.011`, `95% CI 0.993
> to 1.029`), although the confidence interval included the null.

For a categorical exposure, a corresponding sentence might be:

> Compared with `ECOG0`, worse performance status was associated with a higher
> incident hazard, with the strongest contrast observed for `ECOG2`
> (`HR = 2.579`, `95% CI 1.660 to 4.007`) after adjustment for sex.

The scientific style is the same in both cases:

- identify the exposure contrast
- report `HR`, `95% CI`, and model level
- describe the estimate conservatively

# What to take away from Step 1

After this step, you should already know:

1. how to structure a survival-ready cohort data frame
2. how to inspect event burden and follow-up before fitting
3. how to choose between continuous and categorical exposure coding
4. how to define an adjustment hierarchy before fitting
5. how to interpret a main Cox table in manuscript language

Step 2 then asks a different question:

> does the exposure effect differ across groups, and on which scale?

That belongs to subgroup and interaction analysis, not to the main Cox result
itself.
