---
title: "Survival Analysis Step 3.0: Mediation Analysis"
description: "Learn how to choose settings, fit models, and interpret survival mediation analysis in leo.ukb."
output:
  rmarkdown::html_vignette:
    highlight: default
vignette: >
  %\VignetteIndexEntry{Survival Analysis Step 3.0: Mediation Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(leo.ukb))
has_regmedint <- requireNamespace("regmedint", quietly = TRUE)
if (!has_regmedint) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

# What this step is for

This tutorial is the Step 3.0 survival-analysis article in `leo.ukb`. After a
main Cox model and any subgroup or interaction work, a different question may
arise:

> does part of the exposure-outcome association operate through a mediator?

That is a mediation question, not a subgroup question and not an interaction
question.

```{r, echo = FALSE, results = "asis"}
if (!has_regmedint) {
  cat(
    "> Note: regmedint is not installed in the current build environment, ",
    "so the example code below is shown but not executed. Install regmedint ",
    "locally to run the full tutorial.\n\n",
    sep = ""
  )
}
```

# First separate mediation from interaction

The current survival-analysis workflow in `leo.ukb` answers three different
questions:

1. Step 1: is there a main exposure-outcome association?
2. Step 2: does that association differ across groups, and on which scale?
3. Step 3.0: is part of that association transmitted through a mediator?

This distinction matters because:

- interaction is about **effect heterogeneity**
- mediation is about a **pathway decomposition**

In other words, mediation does not replace Step 1 or Step 2. It builds on
them.

# Start from the data and analytic roles

We use a synthetic example with a rare event, a binary exposure, a binary
mediator, and several covariates.

```{r}
set.seed(20260323)
n <- 500
age <- rnorm(n, 60, 8)
bmi <- rnorm(n, 26, 4)
smoking <- factor(sample(c("Never", "Ever"), n, replace = TRUE))
exposure_num <- rbinom(n, 1, plogis(-0.2 + 0.02 * (age - 60)))
mediator_num <- rbinom(n, 1, plogis(-0.5 + 0.9 * exposure_num + 0.02 * (bmi - 26)))
hazard_rate <- exp(
  -5.8 + 0.55 * exposure_num + 0.40 * mediator_num +
    0.02 * (age - 60) + 0.03 * (bmi - 26) + 0.20 * (smoking == "Ever")
)
event_time <- rexp(n, rate = hazard_rate)
censor_time <- rexp(n, rate = 0.08)

med_df <- data.frame(
  outcome = as.integer(event_time <= censor_time),
  outcome_censor = pmax(pmin(event_time, censor_time), 0.2),
  exposure = factor(exposure_num, levels = c(0, 1), labels = c("Low risk", "High risk")),
  mediator = factor(
    mediator_num,
    levels = c(0, 1),
    labels = c("Low inflammation", "High inflammation")
  ),
  age = age,
  bmi = bmi,
  smoking = smoking
)
```

Before fitting anything, look at the analytic roles explicitly.

```{r}
dplyr::glimpse(med_df)
```

Table: Analytic roles in the Step 3 teaching data frame

| Variable | Type | Step 3 role |
| --- | --- | --- |
| `outcome` | Binary event indicator | Incident outcome event |
| `outcome_censor` | Follow-up time | Incident outcome follow-up time |
| `exposure` | Binary exposure | Exposure to be decomposed |
| `mediator` | Binary mediator | Candidate pathway variable |
| `age` | Continuous covariate | Adjustment covariate |
| `bmi` | Continuous covariate | Adjustment covariate |
| `smoking` | Binary covariate | Example covariate that requires explicit `x_cov_cond` |

# Inspect the analysis-ready data before fitting

As in Step 1 and Step 2, it helps to inspect the analysis-ready cohort before
moving to the model.

```{r}
knitr::kable(utils::head(med_df, 6))
```

```{r}
event_summary <- data.frame(
  Metric = c("Participants", "Events", "Event rate"),
  Value = c(
    nrow(med_df),
    sum(med_df$outcome),
    sprintf("%.1f%%", 100 * mean(med_df$outcome))
  )
)

knitr::kable(
  event_summary,
  caption = "Overall event burden in the mediation teaching data"
)
```

```{r}
knitr::kable(
  dplyr::count(med_df, exposure, mediator),
  caption = "Exposure-mediator joint counts"
)
```

These simple checks already answer three important questions:

- is the event rate low enough that `yreg = "auto"` will reasonably choose the
  Cox path?
- does the mediator distribution differ across exposure groups in a direction
  consistent with a possible pathway?
- do the covariates look interpretable enough to define a reproducible
  evaluation setting later?

# Choose the mediation setup before fitting

Mediation analysis is easier to defend when the design choices are explicit
before the model is run.

```{r}
setup_table <- data.frame(
  Component = c(
    "Exposure contrast",
    "Mediator type",
    "Outcome model",
    "Covariate evaluation profile"
  ),
  Current_rule = c(
    "Binary factor by default, or numeric with explicit x_exp_a0 and x_exp_a1",
    "Binary mediator -> logistic; continuous/score-like mediator -> linear",
    "yreg = 'auto' uses survCox when event rate <= 10%, otherwise survAFT_weibull",
    "x_cov_cond should be explicit when x_cov contains non-numeric or binary numeric variables"
  ),
  Why_it_matters = c(
    "All direct and indirect effects are defined relative to this comparison",
    "The mediator model determines how the pathway is represented",
    "The final ratio scale depends on the actual outcome model",
    "The decomposition is always evaluated at a specific covariate profile"
  )
)

knitr::kable(
  setup_table,
  caption = "Choose the mediation setup before fitting"
)
```

# Know the method boundary before fitting

`leo_cox_mediation()` currently wraps `regmedint::regmedint()` and keeps two
model settings parallel:

- `mreg` controls the mediator model, i.e. `x_med ~ x_exp + x_cov`
- `yreg` controls the outcome model, i.e. `time ~ x_exp + x_med + x_cov`
- when `interaction = TRUE`, the outcome model becomes
  `time ~ x_exp + x_med + x_exp:x_med + x_cov`

The current wrapper is most appropriate when:

- the outcome is time-to-event
- the exposure is binary, or numeric with explicit `x_exp_a0` and `x_exp_a1`
- the mediator is binary or continuous
- the temporal ordering is scientifically defensible

Two boundaries are especially important in UK Biobank-style analyses that rely
heavily on baseline measurements:

1. if the final `yreg` (the outcome-model setting here) is `survCox`, the
   decomposition relies on the rare-event approximation; in plain terms, this
   means the split is easier to defend as an approximate result when events are
   relatively rare, rather than as an exact estimate
2. if exposure and mediator are measured at the same baseline visit, the
   data usually cannot show that exposure came before mediator, so the result
   is better written as decomposition evidence that is consistent with an
   assumed pathway; without external evidence on temporal ordering, it should
   not be written up as a strong causal claim

# Fit the simplest mediation workflow

As in the Cox tutorials, define the adjustment hierarchy first.

```{r}
med_models <- list(
  "Model A" = c("age"),
  "Model B" = c("age", "bmi")
)

res_med <- leo_cox_mediation(
  df = med_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "exposure",
  x_med = "mediator",
  x_cov = med_models,
  verbose = FALSE
)

res_med$result
```

# How to read the result table

The mediation output is easier to read if you separate the columns into roles:

- `Effect code` and `Effect` identify which decomposition quantity is being
  shown
- `Scale` reminds you that `pm` is a proportion, whereas the other rows are
  ratio measures
- `Exposure contrast` shows the `x_exp_a0 -> x_exp_a1` comparison
- `Mediator reference` is populated for the `cde` row and echoed in
  `$evaluation`
- `Mediator model` and `Outcome model` show the actual `mreg` and `yreg`
  selected

The scientific point is simple:

> mediation estimates are not free-floating numbers; they are conditional on a
> chosen exposure contrast, mediator reference, covariate profile, and model
> pair

# Teach the core settings in `leo.ukb` language

`regmedint` uses a compact set of evaluation settings and model settings.
`leo_cox_mediation()` keeps the same scientific meaning while presenting them
in more readable `leo.ukb` language.

| Question | `leo.ukb` parameter | What it means in practice |
|---|---|---|
| Which exposure contrast is being compared? | `x_exp_a0`, `x_exp_a1` | The exposure comparison used to define the mediation effect, such as `Low risk -> High risk` or `8.0 -> 9.0`. |
| At what mediator value is the controlled direct effect evaluated? | `x_med_cde` | The mediator reference value used for `cde`. |
| Under what covariate profile are the effects evaluated? | `x_cov_cond` | The covariate setting at which the effects are computed. |
| How is the mediator modeled? | `mreg` | The mediator model for `x_med ~ x_exp + x_cov`. |
| How is the survival outcome modeled? | `yreg` | The outcome model for `time ~ x_exp + x_med + x_cov`. |
| Should the outcome model include `x_exp:x_med`? | `interaction` | If `TRUE`, the outcome model allows the effect of `x_exp` to vary with `x_med`. |

In other words:

- `x_exp_a0`, `x_exp_a1`, `x_med_cde`, and `x_cov_cond` define **where** the
  mediation effects are evaluated
- `mreg`, `yreg`, and `interaction` define **which models** generate that
  decomposition

# How `mreg = "auto"` behaves

The current `leo.ukb` rule is:

- binary factor / character / logical mediator -> `logistic`
- two-valued integer-like numeric mediator such as `0/1` -> `logistic`
- numeric mediator with more than two observed values, or non-integer
  two-valued coding such as `0.2/0.8` -> `linear`
- multi-level categorical mediator stored as factor -> not currently supported

This is the practical interpretation:

- use `auto` freely for binary mediators
- use `auto` for true continuous biomarkers
- for ordered scores, such as a low-grade inflammation index from `-7` to `7`,
  keep the variable numeric and usually let it behave as a continuous mediator
- if the scientific question is threshold-based, dichotomize first and then
  use a logistic mediator model

# Decode the abbreviations before interpreting anything

```{r}
data.frame(
  `Effect code` = c("cde", "pnde", "tnie", "tnde", "pnie", "te", "pm"),
  Meaning = c(
    "Controlled direct effect",
    "Pure natural direct effect",
    "Total natural indirect effect",
    "Total natural direct effect",
    "Pure natural indirect effect",
    "Total effect",
    "Proportion mediated"
  ),
  `Practical reading` = c(
    "Direct effect when the mediator is fixed at the chosen reference value.",
    "Direct pathway not operating through the mediator.",
    "Indirect pathway operating through the mediator.",
    "Direct pathway allowing the mediator to vary naturally.",
    "Indirect pathway under the pure natural definition.",
    "Overall exposure-outcome effect.",
    "Share of the total effect attributed to the mediator pathway."
  )
)
```

For most applied readers, three rows are especially important:

- `te`: how large the overall association is
- `tnie`: how large the mediator pathway appears to be
- `pm`: what fraction of the total effect may be operating through the mediator

# Turn the result into a clinically readable summary

```{r}
key_rows <- dplyr::filter(
  res_med$result,
  Model == "Model B",
  `Effect code` %in% c("te", "tnie", "pm")
)
key_rows

te_row <- dplyr::filter(key_rows, `Effect code` == "te")
tnie_row <- dplyr::filter(key_rows, `Effect code` == "tnie")
pm_row <- dplyr::filter(key_rows, `Effect code` == "pm")
pm_ci <- as.numeric(trimws(strsplit(pm_row$`95% CI`, ",", fixed = TRUE)[[1]]))
ratio_label <- if (te_row$`Outcome model` == "survCox") "HR" else "TR"
ratio_noun <- if (te_row$`Outcome model` == "survCox") "hazard" else "time"
te_pct <- 100 * (te_row$Estimate - 1)
pm_pct <- 100 * pm_row$Estimate

data.frame(
  Quantity = c("Total effect", "Indirect effect through inflammation", "Proportion mediated"),
  `Model B summary` = c(
    sprintf("%s %.3f (%s)", ratio_label, te_row$Estimate, te_row$`95% CI`),
    sprintf("%s %.3f (%s)", ratio_label, tnie_row$Estimate, tnie_row$`95% CI`),
    sprintf("%.1f%% (%.1f%% to %.1f%%)", 100 * pm_row$Estimate, 100 * pm_ci[1], 100 * pm_ci[2])
  )
)
```

Because the event rate in this example is only
`r sprintf("%.1f%%", 100 * mean(med_df$outcome))`, `yreg = "auto"` selected
`survCox`, so the ratio scale is a hazard ratio.

In manuscript-style language, the Model B result can be read as:

- the `High risk` group has an overall `r ratio_noun` ratio that is about
  `r sprintf("%.1f", abs(te_pct))`%
  `r if (te_pct >= 0) "higher" else "lower"` than the `Low risk` group
  (`te = `r sprintf("%.3f", te_row$Estimate)``)
- the inflammation pathway contributes only a modest indirect effect
  (`tnie = `r sprintf("%.3f", tnie_row$Estimate)``)
- the point estimate for the proportion mediated is `r sprintf("%.1f%%", pm_pct)`,
  suggesting that part of the overall association could be operating through
  the inflammation pathway

The critical scientific qualifier is uncertainty. Here the confidence interval
for `pm` is wide and crosses zero, so the careful interpretation is not
"mediation has been established", but:

> the estimates are compatible with partial mediation through inflammation, but
> the pathway effect remains too imprecise to support a strong mechanistic
> claim on its own

# Inspect the evaluation settings directly

`leo_cox_mediation()` keeps a compact evaluation summary in `$evaluation`.

```{r}
res_med$evaluation
```

This table is especially useful when you want to audit:

- which `x_exp_a0 -> x_exp_a1` contrast was used
- which mediator reference was used for the controlled direct effect
- which covariate profile was used for `x_cov_cond`
- which `mreg` and `yreg` were used
- whether the outcome model included `x_exp:x_med`

# Non-numeric covariates need explicit `x_cov_cond`

When the selected adjustment set contains a non-numeric covariate, the
evaluation profile must be given explicitly. The same applies to binary
numeric covariates such as `0/1` indicators, because the default median can
otherwise create impossible values such as `0.5`.

```{r}
res_med_smoking <- leo_cox_mediation(
  df = med_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "exposure",
  x_med = "mediator",
  x_cov = "smoking",
  x_cov_cond = list(smoking = "Never"),
  verbose = FALSE
)

res_med_smoking$evaluation
```

This is not just more transparent. It is required whenever `x_cov` contains a
factor or binary numeric covariate, because the evaluation profile has to map
to a real person that the reader can reconstruct.

# Continuous mediators, including score-like mediators, are also supported

The same workflow can also be used with a continuous mediator.

```{r}
med_df_cont <- dplyr::mutate(
  med_df,
  mediator_cont = rnorm(
    n,
    mean = 0.5 * as.integer(exposure == "High risk") + 0.02 * (age - 60),
    sd = 1
  )
)

res_med_cont <- leo_cox_mediation(
  df = med_df_cont,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "exposure",
  x_med = "mediator_cont",
  x_cov = "age",
  mreg = "linear",
  verbose = FALSE
)

res_med_cont$result
```

For a continuous mediator, the `cde` row shows a numeric mediator reference
because the controlled direct effect is evaluated at `x_med_cde`.

This is also the workflow we would usually recommend for an ordered score-like
mediator. For example, if a low-grade inflammation index ranges from `-7` to
`7`, the first question is scientific rather than computational:

- does the index behave like a severity score, where one-unit changes are
  interpretable enough for a linear approximation?
- or is the real question about a thresholded state that should be defined as
  binary first?

For most ordered severity scores, the cleaner first pass is to keep the score
numeric and fit a linear mediator model.

# Add a paper-style pathway figure

For a single mediator, a compact pathway figure often communicates the result
faster than a dense table alone.

```{r, fig.width = 8.8, fig.height = 5.8, dev.args = list(bg = "white")}
leo_cox_mediation_plot(
  x = res_med,
  model = "Model B",
  exposure_label = "Metabolic\nrisk",
  mediator_label = "Inflammation",
  outcome_label = "Incident\noutcome",
  language = "en",
  palette = "jama"
)
```

This figure is useful because it keeps the main teaching elements in one
place:

- left box: the exposure contrast
- middle box: the proposed mediator
- right box: the incident outcome
- top text: the indirect pathway through the mediator
- middle text: the estimated proportion of the total effect attributed to that
  pathway
- bottom text: the direct pathway not explained by the mediator

One subtle but important point is that the mediated pathway shown here is the
full `X -> M -> Y` pathway. The figure does not print a separate `M -> Y`
coefficient because `regmedint` reports causal decomposition quantities such as
`tnie`, `pnde`, and `te`, rather than a simple product of two regression
coefficients.

# Compare the built-in journal palettes

The same fitted result can be displayed with several built-in journal-style
palettes.

```{r, fig.width = 10, fig.height = 9, fig.show = "hold", out.width = "49%", dev.args = list(bg = "white")}
for (pal in c("jama", "jco", "lancet", "nejm")) {
  leo_cox_mediation_plot(
    x = res_med,
    model = "Model B",
    exposure_label = "Metabolic\nrisk",
    mediator_label = "Inflammation",
    outcome_label = "Incident\noutcome",
    language = "en",
    palette = pal
  )
}
```

# How to interpret common result patterns

In applied survival mediation analysis, the most common interpretation patterns
are:

1. `te` is clearly away from the null, `tnie` is modest, and `pm` is positive
   but imprecise: compatible with partial mediation, but not decisive
2. `te` and `tnie` are both clearly away from the null, and `pm` is precise:
   stronger evidence that the pathway may explain a meaningful share of the
   total effect
3. `te` is weak or crosses the null: `pm` can become unstable and should be
   treated cautiously
4. `yreg` switches to AFT: all non-`pm` effect rows should now be interpreted
   as time ratios rather than hazard ratios

This is why Step 3 is not only about producing a decomposition table. It is
also about learning when the estimates are scientifically interpretable and
when they are still too unstable for a strong claim.

# How to write this in a paper

A defensible reporting order is usually:

1. report the main exposure-outcome association first
2. then present the mediation analysis as a pathway decomposition
3. state the exposure contrast, mediator reference, and adjustment profile
4. report which `yreg` was used; if it was `survCox`, mention the rare-event
   approximation and make clear that this means the decomposition is only being
   reported as an approximate split when events are relatively rare, and if it
   was AFT, remember that the ratio scale is a time ratio
5. if exposure and mediator are measured at the same baseline visit, write the
   result up as decomposition evidence that is consistent with an assumed
   pathway, and say explicitly that the data alone do not establish exposure
   before mediator unless temporal ordering is supported externally

In other words, mediation should usually refine the biological story after the
main Cox result is already established.

# What to take away from Step 3

This step should leave you with six practical rules:

1. mediation asks a pathway question, not a heterogeneity question
2. the data should be checked before fitting, just as in Step 1 and Step 2
3. `x_exp_a0`, `x_exp_a1`, `x_med_cde`, and `x_cov_cond` define where the
   mediation effects are evaluated
4. `mreg` controls `x_med ~ x_exp + x_cov`, `yreg` controls
   `time ~ x_exp + x_med + x_cov`, and `interaction = TRUE` adds
   `x_exp:x_med` to the outcome model
5. `yreg = "auto"` chooses `survCox` for event proportion `<= 10%` (which
   means the decomposition then relies on the rare-event approximation and is
   better read as approximate when events are relatively rare) and
   `survAFT_weibull` otherwise
6. factor mediators should not be forced into `mreg = "linear"`, and the final
   mediation result should always be interpreted with the actual outcome model,
   uncertainty, and the study's temporal ordering in mind
