---
title: "Survival Analysis Step 1: Main Cox Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survival Analysis Step 1: Main Cox Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(leo.ukb))
```

# What this step is for

This tutorial is the first survival-analysis step in `leo.ukb`. The goal is
simple: fit and interpret the main Cox model before moving on to subgroup,
interaction, or mediation questions.

In applied cohort research, it is often helpful to treat the workflow in two
separate stages:

1. estimate the main exposure-outcome association
2. only then ask whether that association differs across groups

This step covers the first stage.

# Start from a survival-ready data frame

A Cox model is easiest to teach, read, and audit when the data frame already
has one row per participant and clearly separated analytic roles:

- an event indicator
- a follow-up time variable
- an exposure
- covariates for adjustment

Here we build a compact example from `survival::lung`.

```{r}
lung_df <- dplyr::transmute(
  survival::lung,
  outcome = as.integer(status == 2),
  outcome_censor = time / 365.25,
  age = age,
  sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
  ecog_group = factor(
    ph.ecog,
    levels = 0:3,
    labels = c("ECOG0", "ECOG1", "ECOG2", "ECOG3")
  )
)
```

Before fitting any model, check the cohort the same way you would in a real
study: confirm structure, event counts, and follow-up.

```{r}
dplyr::glimpse(lung_df)
```

```{r}
knitr::kable(utils::head(lung_df, 6))
```

```{r}
knitr::kable(
  dplyr::count(lung_df, outcome, sex),
  caption = "Event counts by sex"
)
```

```{r}
followup_summary <- summary(lung_df$outcome_censor)
knitr::kable(
  data.frame(
    Statistic = names(followup_summary),
    Value = as.numeric(followup_summary)
  )
)
```

# Define the adjustment hierarchy before fitting

In cohort studies, model labels usually reflect the scientific argument rather
than mere coding convenience. `leo_cox()` accepts `x_cov` as a named list, and
those names are carried directly into the output table.

```{r}
cox_models <- list(
  "Crude" = NULL,
  "Model A" = c("sex"),
  "Model B" = c("sex", "ecog_group")
)

cox_models
```

This small habit keeps the eventual table close to manuscript form from the
start.

# Fit and read the main Cox model

`leo_cox()` is the main entry point for survival analysis in `leo.ukb`.

```{r}
res_cox <- leo_cox(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_cov = cox_models,
  verbose = FALSE
)

res_cox$result
```

The default output is already manuscript-oriented:

- one row per exposure result
- one block of columns per model
- directly formatted `HR`, `95% CI`, and `P value`

The cumulative follow-up column is reported as `Person-years`. In this example,
the original follow-up time was converted from days to years before fitting the
model, so the table matches usual cohort reporting style.

# Use tidy output when you need programmable rows

The wide table is convenient for reading. The tidy table is better when the
next step is plotting, row binding, or programmatic post-processing.

```{r}
dplyr::select(
  leo_cox_format(res_cox, style = "tidy"),
  Model, Exposure, Outcome, HR, `95% CI`, `P value`
)
```

A practical rule is:

- use `$result` for manuscript-style reading
- use `leo_cox_format(..., style = "tidy")` for downstream programming

# The same workflow handles categorical exposures

Many real analyses alternate between continuous and grouped exposures. The
workflow does not change; you simply declare the exposure type explicitly.

```{r}
res_cox_cat <- leo_cox(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "ecog_group",
  x_cov = list("Crude" = NULL, "Model A" = c("sex")),
  x_exp_type = "categorical",
  verbose = FALSE
)

res_cox_cat$result
```

This consistency is useful in practice because continuous and categorical
exposures can still be presented in the same reporting grammar.

# What to take to step 2

After step 1, you should already know:

1. how to structure a survival-ready cohort data frame
2. how to define a model hierarchy before fitting
3. how to run `leo_cox()`
4. how to read the wide and tidy Cox outputs

Step 2 then asks a different question:

> does the exposure effect differ across groups?

That second question belongs to subgroup and interaction analysis, not to the
main Cox table itself.
