---
title: "Survival Analysis Step 1: Main Cox Analysis"
description: "Learn how to prepare, fit, and interpret the main Cox model for an incident survival outcome."
output:
  rmarkdown::html_vignette:
    highlight: default
vignette: >
  %\VignetteIndexEntry{Survival Analysis Step 1: Main Cox Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(leo.ukb))
```

# What this step is for

This tutorial is the first survival-analysis step in `leo.ukb`. Before we talk
about subgroup differences, interaction, or mediation, we need a clean answer
to the first question:

> is there a main exposure-outcome association in the cohort?

That is the role of the main Cox model.

In practice, a strong survival-analysis workflow is easier to defend when it
is split into stages:

1. establish the main exposure-outcome association
2. only then ask whether that association differs across groups
3. only then ask whether part of that association may operate through a
   mediator

Step 1 is therefore the foundation for everything that comes later.

# Start from the cohort question and analytic roles

We use `survival::lung` as a compact teaching example, but we structure the
data frame the same way we would for a real cohort analysis.

```{r}
lung_df <- dplyr::transmute(
  survival::lung,
  outcome = as.integer(status == 2),
  outcome_censor = time / 365.25,
  age = age,
  sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
  ecog_group = factor(
    ph.ecog,
    levels = 0:3,
    labels = c("ECOG0", "ECOG1", "ECOG2", "ECOG3")
  )
)
```

Now check the analytic roles explicitly instead of jumping straight into model
fitting.

Table: Analytic roles in the Step 1 teaching data frame

| Variable | Type | Step 1 role |
| --- | --- | --- |
| `outcome` | Binary event indicator | Incident outcome event |
| `outcome_censor` | Follow-up time in years | Incident outcome follow-up time |
| `age` | Continuous exposure | Main exposure in the first example |
| `sex` | Binary covariate | Basic demographic adjustment |
| `ecog_group` | 4-level ECOG performance-status grouping | ECOG performance-status adjustment |

# Inspect the analysis-ready data before fitting

Main Cox analysis is easier to trust when the cohort structure, event counts,
and follow-up burden have already been checked.

```{r}
dplyr::glimpse(lung_df)
```

```{r}
knitr::kable(utils::head(lung_df, 6))
```

```{r}
event_summary <- data.frame(
  Metric = c("Participants", "Events", "Event rate"),
  Value = c(
    nrow(lung_df),
    sum(lung_df$outcome),
    sprintf("%.1f%%", 100 * mean(lung_df$outcome))
  )
)

knitr::kable(
  event_summary,
  caption = "Overall event burden"
)
```

```{r}
knitr::kable(
  dplyr::count(lung_df, sex, outcome),
  caption = "Event counts by sex"
)
```

```{r}
followup_summary <- summary(lung_df$outcome_censor)
knitr::kable(
  data.frame(
    Statistic = names(followup_summary),
    Value = as.numeric(followup_summary)
  ),
  caption = "Follow-up summary in years"
)
```

These quick checks already tell us whether the event rate, subgroup size, and
follow-up scale look plausible enough for a first-pass Cox analysis.

A careful supervisor would still pause here before touching the hazard ratio.
The first question is whether this looks like a cohort that can support
interpretation at all: are the events too sparse, is the follow-up too thin,
or are some cells already so small that they can only support a directional
reading? This quiet step prevents a surprising amount of over-interpretation
later.

# Choose the exposure scale before fitting

Before fitting anything, decide whether your exposure is being treated as
continuous or categorical. This decision changes what the hazard ratio means.

```{r}
scale_table <- data.frame(
  Exposure_setup = c("Continuous exposure", "Categorical exposure"),
  Typical_question = c(
    "What is the change in hazard per one-unit increase?",
    "How does each category compare with the reference level?"
  ),
  Example_in_this_step = c("age", "ecog_group"),
  Main_setting = c("Default", "x_exp_type = 'categorical'")
)

knitr::kable(
  scale_table,
  caption = "Choose the exposure scale before fitting the main Cox model"
)
```

This is the practical rule:

- keep the exposure continuous if the scientific question is per-unit change
- treat it as categorical if the scientific question is level-versus-reference
  comparison

# Define the adjustment hierarchy before fitting

In cohort studies, model labels should reflect the scientific adjustment
hierarchy rather than coding convenience alone. `leo_cox()` accepts a named
list, and those names are carried directly into the final result table.

```{r}
cox_models <- list(
  "Crude" = NULL,
  "Model A" = c("sex"),
  "Model B" = c("sex", "ecog_group")
)

cox_models
```

This habit makes the output read much more like a manuscript table from the
beginning.

# Main Cox example 1: continuous exposure

We first fit the age-outcome association.

```{r}
res_cox <- leo_cox(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_cov = cox_models,
  verbose = FALSE
)

res_cox
```

# How to read the main Cox result

This output is already close to a manuscript table, but it is worth slowing
down for one extra pass:

- `Case N`, `Control N`, and `Person-years` first tell you that the three
  models are being compared on the same complete-case cohort, not on three
  shifting samples
- `Class` tells you whether the exposure was treated as continuous or
  categorical
- the `HR`, `95% CI`, and `P value` columns are the final layer of evidence,
  not the first one

For a continuous exposure, the hazard ratio is interpreted as the relative
hazard change per one-unit increase in the exposure. Here that means per
one-year increase in age. A value such as `HR = 1.011` should therefore be
read as about a 1.1% relative hazard increase, not as a 1.1 percentage-point
absolute risk increase.

To get the tidy table for downstream programming, set `simplify = "tidy"` or
set `simplify = FALSE` and use `leo_cox_format()`:

```{r}
res_cox_full <- leo_cox(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_cov = cox_models,
  simplify = FALSE,
  verbose = FALSE
)
age_tidy <- leo_cox_format(res_cox_full, style = "tidy")
age_model_b <- dplyr::filter(age_tidy, Model == "Model B")
age_model_b
```

In this teaching example, the Model B estimate is `HR = 1.011`, meaning that
after adjustment for sex and ECOG status, the estimated hazard is about
`r sprintf("%.1f", 100 * (age_model_b$HR - 1))`% higher per one-year increase
in age. If you translate that into a more intuitive contrast, a 10-year age
difference corresponds to about `r sprintf("%.1f", 100 * (age_model_b$HR^10 - 1))`%
relative hazard difference. That translation helps build intuition, but the
model is still a per-year Cox coefficient.

The confidence interval, however, includes 1, so the adjusted evidence for a
main age-outcome association is limited in this small example. A professor
reading this line aloud would not stop at `1.011`; they would immediately add
"directionally positive, but still imprecise."

That is exactly how a main Cox result should be interpreted:

- first the direction and size of the point estimate
- then the uncertainty
- then the scientific caution

Beginners often reverse that order by asking "is it significant?" before they
ask whether the effect size is even meaningful. The stronger habit is the
opposite one: first ask how large the association appears to be, then ask how
stable that estimate is, and only then decide how strong the written claim can
be.

# Use tidy output when you need programmable rows

The wide table is convenient for direct reading. The tidy table is better for
row-binding, plotting, or manuscript automation.

```{r}
dplyr::select(
  age_tidy,
  Model, Exposure, Outcome, HR, `95% CI`, `P value`
)
```

A useful rule is:

- use `simplify = "wide"` (the default) or `simplify = "tidy"` when you only
  need the result table
- use `simplify = FALSE` when you need fitted model objects or `leo_cox_format()`

You can also set `p_fmt = "scientific"` to show precise p-values instead of
`<0.001`.

# Main Cox example 2: categorical exposure

The same workflow also handles categorical exposures.

```{r}
res_cox_cat <- leo_cox(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "ecog_group",
  x_cov = list("Crude" = NULL, "Model A" = c("sex")),
  x_exp_type = "categorical",
  verbose = FALSE
)

res_cox_cat
```

This table is read differently from the age model:

- the `Exposure` column now prints `ECOG0 (Ref)` explicitly, so the reference
  row is not hidden in the formatting
- each other row compares one ECOG level with `ECOG0`
- the `Level` column keeps the original subgroup label, which helps later when
  you filter, merge, or plot results
- the interpretation is therefore level-versus-reference rather than per-unit
  increase

If you were reading this as a classroom exercise, the right order would be:
lock onto the reference group first, then move down row by row asking "how far
is this level from the reference", "how stable is that contrast", and "could
this apparently large estimate still be a sparse-cell exaggeration?" That is
how a table starts training judgment rather than merely showing output.

```{r}
ecog_tidy <- leo_cox(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "ecog_group",
  x_cov = list("Crude" = NULL, "Model A" = c("sex")),
  x_exp_type = "categorical",
  simplify = "tidy",
  verbose = FALSE
)
dplyr::select(
  ecog_tidy,
  Model, Exposure, Level, HR, `95% CI`, `P value`
)
```

```{r}
ecog_plot_df <- dplyr::filter(ecog_tidy, Model == "Model A", Level != "ECOG0")
ecog_ci <- do.call(rbind, strsplit(ecog_plot_df$`95% CI`, ", ", fixed = TRUE))
ecog_plot_df$HR <- as.numeric(ecog_plot_df$HR)
ecog_plot_df$ci_l <- as.numeric(ecog_ci[, 1])
ecog_plot_df$ci_u <- as.numeric(ecog_ci[, 2])
ecog_plot_df$Level <- factor(ecog_plot_df$Level, levels = rev(ecog_plot_df$Level))

ggplot2::ggplot(ecog_plot_df, ggplot2::aes(x = HR, y = Level)) +
  ggplot2::geom_vline(xintercept = 1, linetype = 2, color = "#8F8F8F") +
  ggplot2::geom_segment(
    ggplot2::aes(x = ci_l, xend = ci_u, y = Level, yend = Level),
    color = "#4E79A7",
    linewidth = 0.8
  ) +
  ggplot2::geom_point(size = 2.8, color = "#C44E52") +
  ggplot2::scale_x_log10() +
  ggplot2::labs(
    x = "Hazard ratio (log scale)",
    y = NULL,
    title = "Model A hazard ratios for ECOG levels versus the reference"
  ) +
  ggplot2::theme_minimal(base_size = 12)
```

The table and figure together usually teach faster than either one alone. The
farther a point sits from the vertical null line at 1, the stronger the point
estimate departs from the null; the longer the horizontal segment, the more
uncertainty that estimate carries. That is why `ECOG2` looks both strong and
reasonably stable here, while `ECOG3` looks dramatic but highly fragile.

In this example, the adjusted Model A hazard ratios suggest that worse
performance status is associated with a higher incident hazard:

- `ECOG1` versus `ECOG0`: `HR = 1.519`
- `ECOG2` versus `ECOG0`: `HR = 2.579`
- `ECOG3` versus `ECOG0`: `HR = 7.756`

The last row also reminds us why raw output must be read critically: `ECOG3`
has only one event in this small teaching data set, so its confidence interval
is extremely wide. Sparse cells are a scientific warning, not just a cosmetic
table issue.

# How to interpret common Step 1 result patterns

For the main Cox model, the usual interpretation logic is:

1. if the HR is clearly away from 1 and the confidence interval does not cross
   1, the main association is both directionally clear and statistically
   supported
2. if the HR is away from 1 but the confidence interval crosses 1, the data
   suggest a possible association, but the estimate is still imprecise
3. if the point estimate is close to 1, the effect size may be small; however,
   the strength of statistical evidence still depends on the precision of the
   confidence interval, the sample size, and the scientific context
4. for categorical exposures, always read both the effect size and the cell
   sizes; sparse strata can generate unstable large HRs

Step 1 is therefore not only about “running Cox”. It is about learning how to
separate signal, uncertainty, and sparse-data artifacts.

That is often the real mark of maturity in result reading. When a very large
HR appears, the first instinct should not be admiration. It should be the
question: is this a strong signal, or is this the kind of estimate that gets
stretched by too few events in one stratum?

# How to write Step 1 results in a paper

For a continuous exposure, a concise Results-style sentence could be:

> In the fully adjusted model, each one-year increase in age was associated
> with an estimated 1.1% higher incident hazard (`HR = 1.011`, `95% CI 0.993
> to 1.029`), although the confidence interval included the null.

For a categorical exposure, a corresponding sentence might be:

> Compared with `ECOG0`, worse performance status was associated with a higher
> incident hazard, with the strongest contrast observed for `ECOG2`
> (`HR = 2.579`, `95% CI 1.660 to 4.007`) after adjustment for sex.

The scientific style is the same in both cases:

- identify the exposure contrast
- report `HR`, `95% CI`, and model level
- describe the estimate conservatively

# What to take away from Step 1

After this step, you should already know:

1. how to structure a survival-ready cohort data frame
2. how to inspect event burden and follow-up before fitting
3. how to choose between continuous and categorical exposure coding
4. how to define an adjustment hierarchy before fitting
5. how to interpret a main Cox table in manuscript language

Step 2 then asks a different question:

> does the exposure effect differ across groups, and on which scale?

That belongs to subgroup and interaction analysis, not to the main Cox result
itself.
