---
title: "Survival Analysis with leo.ukb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survival Analysis with leo.ukb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(leo.ukb))
```

# The clinical question

In clinical cohort research, survival analysis usually begins with a simple
question: after follow-up begins, is an exposure associated with the rate at
which an event occurs?

This tutorial shows how to answer that question with `leo.ukb`, using one
coherent workflow:

- build an analysis-ready survival data frame
- define a clear adjustment hierarchy
- fit and interpret the main Cox model
- extend the analysis to interaction, subgroup, and mediation questions

The aim is not to list functions. The aim is to show how a well-structured
survival analysis can move from data preparation to publication-ready tables
without losing statistical clarity.

# Start with a survival-ready data frame

Before fitting any model, it is worth pausing to define the data structure.
A Cox analysis is easiest to manage when the data frame has one row per
participant and one column for each major analytic role:

- an event indicator
- a follow-up time variable
- an exposure
- covariates for adjustment or stratification

Here we build a compact example from `survival::lung`.

```{r}
lung_df <- dplyr::transmute(
  survival::lung,
  outcome = as.integer(status == 2),
  outcome_censor = time,
  age = age,
  sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
  ecog_group = factor(
    ph.ecog,
    levels = 0:3,
    labels = c("ECOG0", "ECOG1", "ECOG2", "ECOG3")
  )
)
```

Before fitting the model, inspect the cohort the way you would in a real study:
first confirm the data structure, then look at event counts and follow-up.

```{r}
dplyr::glimpse(lung_df)
```

```{r}
knitr::kable(utils::head(lung_df, 6))
```

```{r}
knitr::kable(
  dplyr::count(lung_df, outcome, sex),
  caption = "Event counts by sex"
)
```

```{r}
followup_summary <- summary(lung_df$outcome_censor)
knitr::kable(
  data.frame(
    Statistic = names(followup_summary),
    Value = as.numeric(followup_summary)
  )
)
```

This example cohort already contains the columns needed for a basic survival
analysis:

- `outcome`: event indicator
- `outcome_censor`: follow-up time
- `age`: continuous exposure
- `sex` and `ecog_group`: covariates or stratifying variables

Once your own cohort data are arranged in this way, the downstream analysis
becomes much easier to read, audit, and explain.

# Define the adjustment strategy before fitting

In cohort studies, the model hierarchy often reflects the scientific argument:
which variables belong in the crude model, which enter the minimally adjusted
model, and which are reserved for the fuller clinical model.

`leo_cox()` accepts `x_cov` as a named list, and those names are carried
forward to the output table.

```{r}
cox_models <- list(
  "Crude" = NULL,
  "Model A" = c("sex"),
  "Model B" = c("sex", "ecog_group")
)

cox_models
```

This is a small but useful discipline. It means the final table will already be
labeled as `Crude HR`, `Model A HR`, and `Model B HR`, instead of requiring
manual relabeling after the analysis is finished.

# Fit and read the main Cox model

`leo_cox()` is the main entry point for time-to-event analysis in `leo.ukb`.
The key interface is straightforward:

- `df`: the analysis data frame
- `y_out = c(event, time)`: event and follow-up columns
- `x_exp`: the exposure
- `x_cov`: the named list of adjustment models

```{r}
res_cox <- leo_cox(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_cov = cox_models,
  verbose = FALSE
)

res_cox$result
```

The default output is already manuscript-oriented:

- one row per exposure result
- one block of columns per model
- directly formatted `HR`, `95% CI`, and `P value`

In routine cohort analyses, this is often close to the table that ultimately
appears in the main results section. It can therefore serve both as a quick
analytic check and as the starting point for presentation.

The cumulative follow-up column is reported as `Person-time`, meaning the total
participant time in the original unit of the input time variable. In this
teaching example, `survival::lung` records time in days rather than years.

# Use tidy output when you need programmable rows

The wide table is convenient for reading. The tidy table is better when the
next step is plotting, row binding, or programmatic post-processing.

```{r}
dplyr::select(
  leo_cox_format(res_cox, style = "tidy"),
  Model, Exposure, Outcome, HR, `95% CI`, `P value`
)
```

The practical rule is simple:

- use `$result` when you want a manuscript-style summary
- use `leo_cox_format(..., style = "tidy")` when you want analysis-friendly rows

# The same workflow handles categorical exposures

Many applied analyses alternate between continuous and grouped exposures.
The workflow does not change; the only additional instruction is to declare the
exposure type explicitly.

```{r}
res_cox_cat <- leo_cox(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "ecog_group",
  x_cov = list("Crude" = NULL, "Model A" = c("sex")),
  x_exp_type = "categorical",
  verbose = FALSE
)

res_cox_cat$result
```

The useful point here is not simply that the model runs. It is that the
reference row and the level-specific rows are returned in the same table style
as the continuous-exposure analysis, which keeps the reporting grammar
consistent across analyses.

# Interaction analysis: does the exposure effect differ across groups?

Once the main association is estimated, a natural next question is whether the
effect is the same in all groups. That question should not be answered by
visually comparing subgroup-specific `P values`. It should be answered by a
formal interaction model.

`leo_cox_interaction()` reports that formal interaction test.

```{r}
res_inter <- leo_cox_interaction(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_inter = "sex",
  x_cov = list("Crude" = NULL, "Model A" = c("ecog_group")),
  verbose = FALSE
)

res_inter$result
```

When subgroup analyses are reported in a paper, this interaction table is the
formal evidence for effect modification.

# Subgroup analysis: combine stratum-specific HRs with interaction P

After establishing whether interaction is present, it is often useful to show
the stratum-specific hazard ratios themselves. `leo_cox_subgroup()` combines:

- subgroup-specific hazard ratios
- the matching `P for interaction`

```{r}
res_sub <- leo_cox_subgroup(
  df = lung_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "age",
  x_subgroup = c("sex", "ecog_group"),
  x_cov = list("Crude" = NULL),
  verbose = FALSE
)

utils::head(res_sub$result, 8)
```

In most reports, `P for interaction` should be treated as the primary
heterogeneity signal. If `P for heterogeneity` is also shown, it is usually
best regarded as supporting information rather than the main decision point.

# Mediation analysis: from association to pathway-oriented questions

Some studies stop after estimating the exposure-outcome association. Others go
one step further and ask whether part of that association may operate through a
candidate mediator. `leo_cox_mediation()` is designed for that second type of
question.

In its current form, the most reliable first-pass use cases are:

- binary exposure
- continuous or binary mediator
- mostly numeric covariates

Because `survival::lung` was not designed as a mediation example, it is clearer
to use a compact synthetic teaching dataset for this section.

```{r}
if (requireNamespace("regmedint", quietly = TRUE)) {
  set.seed(123)
  n <- 300

  med_df <- data.frame(
    outcome = rbinom(n, 1, plogis(-0.4 + 0.7 * rbinom(n, 1, 0.5))),
    outcome_censor = rexp(n, rate = 0.08) + 0.1,
    exposure = factor(rbinom(n, 1, 0.5), levels = c(0, 1), labels = c("No", "Yes")),
    mediator = factor(rbinom(n, 1, 0.45), levels = c(0, 1), labels = c("Low", "High")),
    age = rnorm(n, 60, 8),
    bmi = rnorm(n, 26, 4)
  )

  dplyr::glimpse(med_df)
  knitr::kable(utils::head(med_df, 6))
}
```

```{r}
if (requireNamespace("regmedint", quietly = TRUE)) {
  res_med <- leo_cox_mediation(
    df = med_df,
    y_out = c("outcome", "outcome_censor"),
    x_exp = "exposure",
    x_med = "mediator",
    x_cov = list("Model A" = c("age"), "Model B" = c("age", "bmi")),
    verbose = FALSE
  )

  res_med$result
}
```

For a first reading, focus on these columns:

- `Effect`: which causal quantity is being reported
- `Estimate`: the point estimate
- `95% CI`: uncertainty interval
- `P value`: statistical support

The exact interpretation of `cde`, `pnde`, `tnie`, `te`, and `pm` depends on
the mediation estimand. For a first pass, the important lesson is more modest:
understand what the input data should look like and how the results are
organized when a mediation analysis is appropriate.

# Take-home summary

The survival-analysis functions in `leo.ukb` are easiest to learn when treated
as one connected workflow rather than a list of unrelated helpers:

- `leo_cox()` estimates the main exposure-outcome association
- `leo_cox_interaction()` tests whether that association differs across groups
- `leo_cox_subgroup()` displays stratum-specific HRs together with interaction P
- `leo_cox_mediation()` extends the analysis toward pathway-oriented questions

The workflow therefore grows outward from the same core:

1. prepare a survival-ready data frame
2. define the model hierarchy once
3. fit the main Cox model
4. expand to interaction, subgroup, or mediation only when the scientific
   question changes

If these four steps are clear, the package becomes much easier to use in a real
cohort study.
