---
title: "Survival Analysis Step 3: Mediation Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survival Analysis Step 3: Mediation Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(leo.ukb))
has_regmedint <- requireNamespace("regmedint", quietly = TRUE)
if (!has_regmedint) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

# What this step is for

This tutorial is the third survival-analysis step in `leo.ukb`. After the main
Cox model and the subgroup/interaction stage, a different question can appear:

> does part of the exposure-outcome association operate through a mediator?

That is a mediation question, not a subgroup question.

```{r, echo = FALSE, results = "asis"}
if (!has_regmedint) {
  cat(
    "> Note: regmedint is not installed in the current build environment, ",
    "so the example code below is shown but not executed. Install regmedint ",
    "locally to run the full tutorial.\n\n",
    sep = ""
  )
}
```

# First separate mediation from interaction

The three steps in the current survival workflow answer three different
questions:

1. Step 1: is there a main exposure-outcome association?
2. Step 2: does that association differ across groups?
3. Step 3: is part of that association transmitted through a mediator?

This distinction matters because mediation is about **pathways**, whereas
interaction is about **effect heterogeneity**.

# Know the method boundary before fitting

`leo_cox_mediation()` wraps `regmedint::regmedint()` with `yreg = "survCox"`.
The current package workflow is most appropriate when:

- the outcome is time-to-event
- the exposure is binary, or numeric with explicit `a0` and `a1`
- the mediator is binary or continuous
- the event is reasonably rare
- the temporal ordering is scientifically defensible

Two practical cautions are especially important in UK Biobank-style cohort work:

1. `regmedint` with `survCox` relies on the rare-event approximation
2. if exposure and mediator are both measured at the same baseline visit, the
   result is easier to defend as a pathway-consistent decomposition than as a
   strong causal claim unless temporal ordering is otherwise justified

# Build a compact teaching data set

We use a small synthetic example with a rare event, a binary exposure, a binary
mediator, and a few covariates.

```{r}
set.seed(20260323)
n <- 500
age <- rnorm(n, 60, 8)
bmi <- rnorm(n, 26, 4)
smoking <- factor(sample(c("Never", "Ever"), n, replace = TRUE))
exposure_num <- rbinom(n, 1, plogis(-0.2 + 0.02 * (age - 60)))
mediator_num <- rbinom(n, 1, plogis(-0.5 + 0.9 * exposure_num + 0.02 * (bmi - 26)))
event_prob <- plogis(-3.1 + 0.55 * exposure_num + 0.40 * mediator_num + 0.02 * (age - 60))

med_df <- data.frame(
  outcome = rbinom(n, 1, event_prob),
  outcome_censor = rexp(n, rate = 0.08) + 0.2,
  exposure = factor(exposure_num, levels = c(0, 1), labels = c("Low risk", "High risk")),
  mediator = factor(mediator_num, levels = c(0, 1), labels = c("Low inflammation", "High inflammation")),
  age = age,
  bmi = bmi,
  smoking = smoking
)

mean(med_df$outcome)
```

The observed event proportion is low enough for this toy example to be aligned
with the intended `survCox` use case.

# Inspect the analysis-ready data before fitting

As in Step 1 and Step 2, it is worth looking at the analysis-ready data first.

```{r}
head(med_df)

dplyr::count(med_df, exposure, mediator)

data.frame(
  metric = c("Participants", "Events", "Event rate"),
  value = c(nrow(med_df), sum(med_df$outcome), sprintf("%.1f%%", 100 * mean(med_df$outcome)))
)
```

This quick check already tells us three useful things:

- the sample size is 500, with 46 observed events
- the event rate is 9.2%, which is low enough for a teaching example built
  around `survCox`
- the `High risk` group has more participants in the `High inflammation`
  category, which is exactly the kind of pathway pattern mediation analysis is
  meant to quantify

# Fit the simplest mediation workflow

As in the Cox tutorials, it helps to define the adjustment hierarchy first.

```{r}
med_models <- list(
  "Model A" = c("age"),
  "Model B" = c("age", "bmi")
)

res_med <- leo_cox_mediation(
  df = med_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "exposure",
  x_med = "mediator",
  x_cov = med_models,
  verbose = FALSE
)

res_med$result
```

# How to read the result table

The mediation output is already organized in a paper-friendly way.

- `Effect code` gives the standard mediation abbreviations such as `cde`,
  `pnde`, `tnie`, and `pm`
- `Effect` gives the full label
- `Scale` reminds you that `pm` is a proportion, while the other mediation
  effects are reported on the hazard-ratio scale
- `Exposure contrast` and `Mediator reference` show exactly where the effects
  were evaluated

This last point is important: mediation estimates are conditional on the chosen
evaluation settings.

# Teach the evaluation settings in `leo.ukb` language

`regmedint` uses four core evaluation arguments, and `leo_cox_mediation()`
keeps the same scientific meaning while presenting them in a more readable way.

| Question | `leo.ukb` parameter | What it means in practice |
|---|---|---|
| Which exposure contrast is being compared? | `a0`, `a1` | The exposure comparison used to define the mediation effect, such as `Low risk -> High risk` or `8.0 -> 9.0`. |
| At what mediator value is the controlled direct effect evaluated? | `m_cde` | The mediator reference value used for `cde`. |
| Under what covariate profile are the effects evaluated? | `c_cond` | The covariate setting at which the effects are computed. |
| Which mediator submodel is used? | `mediator_model` | Whether the mediator is modeled as binary (`logistic`) or continuous (`linear`). |

In other words, `a0`, `a1`, `m_cde`, and `c_cond` are not nuisance options.
They define **where** the direct and indirect effects are evaluated.

# How `mediator_model = "auto"` behaves

The current `leo.ukb` rule is:

- binary factor / character / logical mediator -> `logistic`
- two-valued numeric mediator -> `logistic`
- numeric mediator with more than two observed values -> `linear`
- multi-level categorical mediator stored as factor -> not currently supported

This is usually the safest workflow for applied cohort studies:

- if the mediator is naturally binary, let `auto` pick `logistic`
- if the mediator is a true continuous biomarker, let `auto` pick `linear`
- if the mediator is an ordered score, such as a low-grade inflammation index
  ranging from `-7` to `7`, keep it numeric and treat it as a continuous
  mediator, usually with `mediator_model = "linear"`
- if the scientific question is threshold-based, recode the mediator into a
  clinically justified binary variable first, then use `logistic`

What is **not** recommended is to pass a multi-level categorical mediator as a
factor and expect the current survival mediation wrapper to fit an ordinal or
multinomial mediator model. That is outside the boundary of the present
function.

# Decode the abbreviations before interpreting anything

```{r}
data.frame(
  `Effect code` = c("cde", "pnde", "tnie", "tnde", "pnie", "te", "pm"),
  Meaning = c(
    "Controlled direct effect",
    "Pure natural direct effect",
    "Total natural indirect effect",
    "Total natural direct effect",
    "Pure natural indirect effect",
    "Total effect",
    "Proportion mediated"
  ),
  `Practical reading` = c(
    "Direct effect when the mediator is fixed at the chosen reference value.",
    "Direct pathway not operating through the mediator.",
    "Indirect pathway operating through the mediator.",
    "Direct pathway allowing the mediator to vary naturally.",
    "Indirect pathway under the pure natural definition.",
    "Overall exposure-outcome effect.",
    "Share of the total effect attributed to the mediator pathway."
  )
)
```

For most applied readers, three rows are the most immediately useful:

- `te`: how large the overall association is
- `tnie`: how large the mediator pathway looks on the hazard-ratio scale
- `pm`: what fraction of the total effect may be operating through the mediator

# Turn the result into a clinically readable summary

```{r}
key_rows <- dplyr::filter(res_med$result, Model == "Model B", `Effect code` %in% c("te", "tnie", "pm"))
key_rows

te_row <- dplyr::filter(key_rows, `Effect code` == "te")
tnie_row <- dplyr::filter(key_rows, `Effect code` == "tnie")
pm_row <- dplyr::filter(key_rows, `Effect code` == "pm")
pm_ci <- as.numeric(trimws(strsplit(pm_row$`95% CI`, ",", fixed = TRUE)[[1]]))

data.frame(
  Quantity = c("Total effect", "Indirect effect through inflammation", "Proportion mediated"),
  `Model B summary` = c(
    sprintf("HR %.3f (%s)", te_row$Estimate, te_row$`95% CI`),
    sprintf("HR %.3f (%s)", tnie_row$Estimate, tnie_row$`95% CI`),
    sprintf("%.1f%% (%.1f%% to %.1f%%)", 100 * pm_row$Estimate, 100 * pm_ci[1], 100 * pm_ci[2])
  )
)
```

In this synthetic example, a paper-style reading would be:

- after adjustment for age and BMI, the `High risk` group shows an overall
  hazard that is about 23.4% higher than the `Low risk` group (`te = 1.234`)
- the inflammation pathway contributes only a modest indirect effect
  (`tnie = 1.019`), corresponding to a small upward shift on the hazard-ratio
  scale
- the point estimate for the proportion mediated is 9.7%, which suggests that
  roughly one tenth of the total association could be operating through the
  inflammation pathway in this toy example

The uncertainty, however, is still substantial. Because the confidence interval
for `pm` is wide and crosses zero, the scientifically careful interpretation is
not “mediation has been established”, but:

> the estimates are compatible with partial mediation through inflammation, yet
> they remain too imprecise to support a strong mechanistic claim on their own

If you wanted to write this as a concise Results-style sentence, one reasonable
version would be:

> After adjustment for age and BMI, `High risk` was associated with a higher
> incident hazard (`te = 1.234`), while the indirect pathway through
> inflammation was modest (`tnie = 1.019`). The estimated proportion mediated
> was 9.7%, suggesting possible but uncertain partial mediation through
> inflammation.

# Add a paper-style pathway figure

For a single mediator, a compact path diagram often communicates the result more
clearly than a dense table alone.

```{r, fig.width = 8.8, fig.height = 5.8, dev.args = list(bg = "white")}
leo_cox_mediation_plot(
  x = res_med,
  model = "Model B",
  exposure_label = "Metabolic\nrisk",
  mediator_label = "Inflammation",
  outcome_label = "Incident\noutcome",
  language = "en",
  palette = "jama"
)
```

This kind of figure is useful when writing a paper because it turns the main
mediation quantities into one visual story:

- left box: the exposure contrast being compared
- middle box: the proposed mediator
- right box: the incident outcome
- top text: the indirect pathway through the mediator
- middle text: the estimated proportion of the total effect attributed to that
  pathway
- bottom text: the direct pathway not explained by the mediator

One subtle but important point is that the mediated pathway here is the full
`X -> M -> Y` pathway. The figure does not print a separate `M -> Y`
coefficient because `regmedint` reports causal decomposition quantities such as
`tnie`, `pnde`, and `te`, rather than a simple product of two regression
coefficients.

# Inspect the evaluation settings directly

`leo_cox_mediation()` now keeps a compact evaluation summary in `$evaluation`.

```{r}
res_med$evaluation
```

This table is especially useful when you want to audit:

- which `a0 -> a1` contrast was used
- which mediator reference value was used for the controlled direct effect
- which covariate profile was used for `c_cond`

For example, if your output shows:

- `Exposure contrast = Low risk -> High risk`
- `Mediator reference = High inflammation`
- `c_cond = age=60.000; bmi=26.000`

then the reported direct and indirect effects should be read as:

> the mediation decomposition comparing `High risk` with `Low risk`, evaluated
> at the chosen mediator reference and at that specific covariate profile

# Non-numeric covariates need explicit `c_cond`

When the selected adjustment set contains a non-numeric covariate, the
evaluation profile should be given explicitly.

```{r}
res_med_smoking <- leo_cox_mediation(
  df = med_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "exposure",
  x_med = "mediator",
  x_cov = "smoking",
  c_cond = list(smoking = "Never"),
  verbose = FALSE
)

res_med_smoking$evaluation
```

This is more transparent than letting a factor be silently converted in a way
the reader cannot easily reconstruct.

# Continuous mediators, including score-like mediators, are also supported

The same workflow can be used with a continuous mediator. A truly continuous
mediator, or an ordered score that you are willing to treat as approximately
continuous, should usually be handled with `mediator_model = "linear"`.

```{r}
med_df_cont <- dplyr::mutate(
  med_df,
  mediator_cont = rnorm(n, mean = 0.5 * as.integer(exposure == "High risk") + 0.02 * (age - 60), sd = 1)
)

res_med_cont <- leo_cox_mediation(
  df = med_df_cont,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "exposure",
  x_med = "mediator_cont",
  x_cov = "age",
  mediator_model = "linear",
  verbose = FALSE
)

res_med_cont$result
```

For a continuous mediator, the reference value in the table is numeric because
the controlled direct effect is evaluated at `m_cde`.

This is also the pattern we would usually recommend for an ordered score-like
mediator. For example, if a low-grade inflammation index ranges from `-7` to
`7`, the first question is scientific rather than computational:

- does this index behave like a severity score, where one-unit changes are
  interpretable enough for a linear approximation?
- or is there a clinically meaningful threshold that should define a binary
  mediator instead?

For most ordered severity scores, the cleaner first pass is to keep the score
numeric and fit a linear mediator model. Only if the study question is really
about a thresholded state should you dichotomize it and move to a logistic
mediator model.

# How to write this in a paper

A defensible reporting order is usually:

1. report the main exposure-outcome association first
2. then present the mediation analysis as a pathway decomposition
3. state the exposure contrast, mediator reference, and adjustment profile
4. mention that `survCox` mediation relies on the rare-event approximation
5. avoid over-claiming causality when exposure and mediator are measured at the
   same baseline visit

In other words, mediation should usually refine the biological story after the
main Cox result is already established.

# What to take away from step 3

This step should leave you with four practical rules:

1. mediation asks a pathway question, not a heterogeneity question
2. `a0`, `a1`, `m_cde`, and `c_cond` define where the mediation effects are
   evaluated
3. `leo_cox_mediation()` handles both binary and continuous mediators, but
   factor mediators should not be forced into `mediator_model = "linear"`
4. the `survCox` mediation result should always be interpreted with the
   rare-event assumption and the study's temporal ordering in mind
