---
title: "Survival Analysis Step 3.0: Mediation Analysis"
description: "Learn the default mediation workflow in leo.ukb, from design choices to ACME/ADE/ATE interpretation for survival outcomes."
output:
  rmarkdown::html_vignette:
    highlight: default
vignette: >
  %\VignetteIndexEntry{Survival Analysis Step 3.0: Mediation Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(leo.ukb))
has_mediation <- requireNamespace("mediation", quietly = TRUE)
if (!has_mediation) {
  knitr::opts_chunk$set(eval = FALSE)
}
fmt_num <- function(x, digits = 3) {
  if (length(x) == 0 || all(is.na(x))) return("NA")
  sprintf(paste0("%.", digits, "f"), x[1])
}
fmt_pct <- function(x, digits = 1) {
  if (length(x) == 0 || all(is.na(x))) return("NA")
  sprintf(paste0("%.", digits, "f%%"), 100 * x[1])
}
effect_row <- function(res, code) {
  if (is.null(res)) return(data.frame())
  dplyr::filter(res$result, `Effect code` == code)
}
effect_value <- function(res, code, column = "Estimate") {
  row <- effect_row(res, code)
  if (!nrow(row) || !column %in% names(row)) return(NA_real_)
  row[[column]][1]
}
effect_text <- function(res, code, column = "95% CI") {
  row <- effect_row(res, code)
  if (!nrow(row) || !column %in% names(row)) return("NA")
  as.character(row[[column]][1])
}
eval_value <- function(res, column) {
  if (is.null(res) || !column %in% names(res$evaluation)) return(NA)
  res$evaluation[[column]][1]
}
res_med <- NULL
res_med_profile <- NULL
res_med_boot <- NULL
res_med_inter <- NULL
mediation_warning <- NA_character_
```

# What this step is for

This article is the default Step 3 survival-mediation tutorial in `leo.ukb`.
By the time we arrive here, the main association has usually been established
already, and sometimes Step 2 interaction work has been done as well. The next
question is more mechanistic:

> how much of the exposure-outcome association is compatible with operating
> through a mediator?

That sounds simple, but there are really two tasks hidden inside it:

1. choose a mediation engine that matches the scientific question
2. read the output on the right scale instead of translating everything back
   into a hazard-ratio story by habit

This Step 3.0 tutorial starts with the package default, which is now
`method = "mediation"`.

```{r, echo = FALSE, results = "asis"}
if (!has_mediation) {
  cat(
    "> Note: package `mediation` is not installed in the current build environment, ",
    "so the code is shown but not executed. Install `mediation` locally to run ",
    "the full tutorial.\n\n",
    sep = ""
  )
}
```

# First separate mediation from interaction

Step 2 asked whether an association differed across groups. Step 3 asks
whether one association can be decomposed into pathway pieces. Those are not
the same scientific question.

```{r}
decision_table <- data.frame(
  Step = c("Step 2", "Step 3.0"),
  Main_question = c(
    "Does the exposure-outcome association differ across strata?",
    "Is part of the exposure-outcome association compatible with operating through a mediator?"
  ),
  Typical_output = c(
    "Interaction P value, subgroup HRs, additive-interaction measures",
    "ACME, ADE, ATE, proportion mediated"
  ),
  Core_idea = c(
    "Effect heterogeneity",
    "Pathway decomposition"
  ),
  check.names = FALSE
)

knitr::kable(
  decision_table,
  caption = "Step 2 and Step 3 answer different questions"
)
```

That distinction matters because a mediation result is not a dressed-up
interaction test. Once the two ideas are mixed, the interpretation usually
starts drifting.

A seasoned supervisor often asks a different first question from a beginner.
Not "is it significant?", but "are you talking about heterogeneity or pathway
decomposition?" If that anchor is missing, even a neat ACME/ADE table quickly
loses its interpretive coordinates.

# Build one teaching cohort and inspect the roles first

We will use a synthetic cohort with:

- a binary exposure
- a binary mediator
- two numeric covariates for the first model
- one categorical covariate reserved for the `x_cov_cond` section

The survival outcome is simulated from event time and censoring time, not from
a simple Bernoulli draw, so the tutorial stays aligned with time-to-event data.

```{r}
set.seed(20260329)
n <- 500
age <- rnorm(n, 60, 8)
bmi <- rnorm(n, 26, 4)
smoking <- factor(sample(c("Never", "Ever"), n, replace = TRUE))
exposure_num <- rbinom(n, 1, plogis(-0.2 + 0.02 * (age - 60)))
mediator_num <- rbinom(n, 1, plogis(-0.5 + 0.9 * exposure_num + 0.02 * (bmi - 26)))
hazard_rate <- exp(
  -5.8 + 0.55 * exposure_num + 0.40 * mediator_num +
    0.02 * (age - 60) + 0.03 * (bmi - 26) + 0.20 * (smoking == "Ever")
)
event_time <- rexp(n, rate = hazard_rate)
censor_time <- rexp(n, rate = 0.08)

med_df <- data.frame(
  outcome = as.integer(event_time <= censor_time),
  outcome_censor = pmax(pmin(event_time, censor_time), 0.2),
  exposure = factor(exposure_num, levels = c(0, 1), labels = c("Low risk", "High risk")),
  mediator = factor(
    mediator_num,
    levels = c(0, 1),
    labels = c("Low inflammation", "High inflammation")
  ),
  age = age,
  bmi = bmi,
  smoking = smoking
)
```

```{r}
dplyr::glimpse(med_df)
```

Table: Analytic roles in the Step 3.0 teaching data frame

| Variable | Type | Role in Step 3.0 |
| --- | --- | --- |
| `outcome` | Binary event indicator | Incident event |
| `outcome_censor` | Follow-up time | Event-time outcome |
| `exposure` | Binary factor | Exposure to decompose |
| `mediator` | Binary factor | Candidate pathway variable |
| `age` | Continuous covariate | Numeric adjustment covariate |
| `bmi` | Continuous covariate | Numeric adjustment covariate |
| `smoking` | Binary factor | Teaching example for explicit `x_cov_cond` |

Before fitting the model, look at the data the way you would in a methods
meeting: not just "does the code run?", but "does the analytic story make
sense?"

```{r}
knitr::kable(utils::head(med_df, 6))
```

```{r}
event_summary <- data.frame(
  Metric = c("Participants", "Events", "Event rate"),
  Value = c(
    nrow(med_df),
    sum(med_df$outcome),
    sprintf("%.1f%%", 100 * mean(med_df$outcome))
  )
)

knitr::kable(
  event_summary,
  caption = "Overall event burden in the mediation teaching data"
)
```

```{r}
knitr::kable(
  dplyr::count(med_df, exposure, mediator),
  caption = "Exposure-mediator joint counts"
)
```

Three quick lessons come from these tables:

- the outcome really is time-to-event, not merely binary incidence
- the mediator differs by exposure in a direction that makes a pathway story
  plausible
- the event rate is low here, but the default mediation route still does not
  become a Cox decomposition; it remains a `survreg`-based mediation analysis

# Know what the default route is actually fitting

`leo_cox_mediation()` now defaults to `method = "mediation"`. For survival
outcomes, that means:

```{r}
default_route <- data.frame(
  Component = c(
    "Mediation engine",
    "Outcome model",
    "Scale for non-PM rows",
    "Interaction default",
    "Effect family"
  ),
  Current_setting = c(
    "`mediation::mediate()`",
    "`survival::survreg(dist = 'weibull')`",
    "Time difference",
    "`interaction = NULL` resolves to `FALSE`",
    "ACME / ADE / ATE / PM"
  ),
  Why_it_matters = c(
    "This route reports average mediation effects rather than TE/TNIE/PNDE",
    "The analysis is a parametric survival mediation model, not a Cox mediation model",
    "Negative values imply shorter event-free time, positive values imply longer event-free time",
    "The wrapper stays on the simpler average-effect path unless you explicitly override it",
    "These are the names you should write and interpret in Step 3.0"
  ),
  check.names = FALSE
)

knitr::kable(
  default_route,
  caption = "What the default mediation route means in practice"
)
```

One subtle point deserves emphasis. The internal survival model is currently
`survival::survreg(dist = "weibull")` because that is the supported and stable
route for `method = "mediation"` inside the wrapper. That is a modeling choice,
not a blanket claim that every UK Biobank endpoint is exactly Weibull-shaped.

# Fit the first model with the package default

We start with the simplest defensible version: numeric covariates only,
default engine, default interaction handling.

```{r}
res_med <- leo_cox_mediation(
  df = med_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "exposure",
  x_med = "mediator",
  x_cov = c("age", "bmi"),
  verbose = FALSE
)

res_med$result
```

```{r}
dplyr::select(
  res_med$result,
  `Effect code`, Scale, `Exposure contrast`, Estimate, `95% CI`, `P value`
)
```

```{r}
res_med$evaluation[, c(
  "method", "mreg", "yreg", "outcome_distribution",
  "interaction", "x_cov_cond", "mediation_boot", "mediation_sims"
)]
```

This small evaluation table is worth reading before the effect table:

- `method = mediation` confirms that the default route was used
- `yreg = survreg` tells you the outcome model class that was actually fit
- `outcome_distribution = weibull` tells you which `survreg` distribution was
  used
- `interaction = FALSE` shows the engine-specific default after
  `interaction = NULL`
- `x_cov_cond = NA` is expected in this default run, because the average-effect
  `mediation` route does not force you to pin the decomposition to an explicit
  named profile unless you choose to do so

Learning to read this evaluation table is really about setting rules for the
rest of the interpretation. A good teacher would insist that you state the
engine, the scale, and the presence or absence of an explicit profile before
you say anything about the estimates. Otherwise the same number gets narrated
as a time difference one day and a hazard ratio the next.

# How to read ACME, ADE, ATE, and PM

The first pass should be simple:

- `ACME` is the average indirect effect
- `ADE` is the average direct effect
- `ATE` is the average total effect
- `PM` is the proportion mediated

The second pass is the part people skip too quickly: here the non-PM rows are
on a **time-difference** scale, not a hazard-ratio scale.

```{r}
acme_row <- effect_row(res_med, "acme")
ade_row <- effect_row(res_med, "ade")
ate_row <- effect_row(res_med, "ate")
pm_row <- effect_row(res_med, "pm")

interpret_table <- data.frame(
  Quantity = c("ACME", "ADE", "ATE", "PM"),
  Estimate = c(
    acme_row$Estimate[1],
    ade_row$Estimate[1],
    ate_row$Estimate[1],
    pm_row$Estimate[1]
  ),
  `95% CI` = c(
    acme_row$`95% CI`[1],
    ade_row$`95% CI`[1],
    ate_row$`95% CI`[1],
    pm_row$`95% CI`[1]
  ),
  `How to read it` = c(
    "Average mediated component on the time scale",
    "Average direct component on the time scale",
    "Average overall exposure contrast on the time scale",
    "Share of the total effect carried by the mediator"
  ),
  check.names = FALSE
)

knitr::kable(
  interpret_table,
  caption = "The Step 3.0 effect family"
)
```

The safest reading order is the one a good supervisor would use at the
whiteboard:

1. start with `Effect code` and `Scale`, so you know what family of quantity
   the row is reporting
2. check `Exposure contrast`, so you know that every row is talking about the
   same comparison
3. only then move to `Estimate`, `95% CI`, and `P value`

That order matters because the most common mediation error is not a failed
model fit. It is a failed reading. Once the scale and the exposure contrast
have been stated first, the estimate has a location. Without that, even a very
clean ACME can still be translated into the wrong scientific language.

In this run, the ACME is `r fmt_num(effect_value(res_med, "acme"))`, the ADE is
`r fmt_num(effect_value(res_med, "ade"))`, and the ATE is
`r fmt_num(effect_value(res_med, "ate"))`. Because the outcome model is
`survreg`, these numbers describe differences on the survival-time scale:

- a negative estimate is compatible with shorter event-free time
- a positive estimate is compatible with longer event-free time

That is why it is a mistake to glance at the sign and immediately translate it
into a "higher HR" or "lower HR" story. This default route is not reporting
hazard ratios.

The same result table also carries `Case N`, `Non-event N`, and `Total follow-up`.
Those columns are there to keep the analysis grounded: path decomposition is
still only as trustworthy as the amount of event and follow-up information
supporting it.

The proportion mediated in this run is `r fmt_pct(effect_value(res_med, "pm"))`
with a 95% CI of `r effect_text(res_med, "pm")`. The wrapper also checks
whether the total-effect interval crosses the null. Here
`r if (isTRUE(eval_value(res_med, "pm_unstable"))) "it does, so the proportion mediated is flagged as unstable and should be treated cautiously." else "it does not, so the PM is not automatically flagged as unstable."`

# Name the evaluation profile explicitly when you need it

The default `method = "mediation"` route does not manufacture a named
covariate profile for you. When `x_cov_cond` is omitted, the reported
ACME/ADE/ATE/PM are averaged over the observed covariate distribution. If you
want the decomposition anchored to a specific profile, say so explicitly.

That becomes especially useful once a factor enters `x_cov`, because
"adjusted for smoking" and "evaluated at age 60, smoking = Never" are not the
same scientific claim.

```{r}
res_med_profile <- leo_cox_mediation(
  df = med_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "exposure",
  x_med = "mediator",
  x_cov = c("age", "smoking"),
  x_cov_cond = list(age = 60, smoking = "Never"),
  verbose = FALSE
)

res_med_profile$evaluation[, c("x_cov_cond", "interaction", "method", "yreg")]
```

```{r}
dplyr::select(
  res_med_profile$result,
  `Effect code`, `Covariate profile`, Estimate, `95% CI`
)
```

This matters scientifically, not just technically. Once you add
`smoking = "Never"` to `x_cov_cond`, the mediation effects are being reported
for that covariate profile. That is usually much easier to defend in writing
than a vague statement such as "adjusted for smoking."

This is one of the details teachers usually care about most. "Adjusted for"
only tells the reader that a variable entered the model. It does not, by
itself, tell them on whom the effect is being evaluated. Making `x_cov_cond`
explicit is how statistical adjustment is translated back into a recognizable
human profile.

# Bootstrap is an option, not the only route

`mediation::mediate()` does not require bootstrap in every run. The wrapper
therefore exposes bootstrap controls instead of hard-coding one inference path.
For a quick teaching run, we keep the resampling count modest.

```{r}
res_med_boot <- leo_cox_mediation(
  df = med_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "exposure",
  x_med = "mediator",
  x_cov = c("age", "bmi"),
  mediation_boot = TRUE,
  mediation_sims = 80,
  mediation_boot_ci_type = "perc",
  verbose = FALSE
)

res_med_boot$evaluation[, c(
  "mediation_boot", "mediation_sims", "mediation_boot_ci_type", "mediation_use_speed"
)]
```

The teaching point is not that bootstrap is "more serious" by default. The
teaching point is that inference settings should match the stage of the work:

- a quick exploratory pass may stay with the default simulation-based route
- a final results table may justify bootstrap and a larger `mediation_sims`
  value
- either way, the chosen setting should be visible in the analysis record

# What happens if you force interaction = TRUE?

In the default mediation route, `interaction = NULL` resolves to `FALSE`.
You can still force `TRUE`, but the wrapper warns you for a reason: the
underlying `mediation` object can report treatment-specific quantities, whereas
the current `leo_cox_mediation()` summary still reports averaged
ACME/ADE/ATE/PM.

```{r}
mediation_warning <- NA_character_
res_med_inter <- withCallingHandlers(
  leo_cox_mediation(
    df = med_df,
    y_out = c("outcome", "outcome_censor"),
    x_exp = "exposure",
    x_med = "mediator",
    x_cov = c("age", "bmi"),
    interaction = TRUE,
    verbose = FALSE
  ),
  warning = function(w) {
    mediation_warning <<- conditionMessage(w)
    invokeRestart("muffleWarning")
  }
)

mediation_warning
```

The message is deliberate. If you truly need the treatment-specific mediation
quantities from `mediation::mediate()`, inspect the backend object directly in
`$fit$model_1$mediation_model`, rather than pretending that the four-row
wrapper summary has already become a fully interaction-aware report.

# Draw the pathway figure

The plot wrapper is meant to give you a compact paper-style map of the current
result.

```{r}
leo_cox_mediation_plot(
  x = res_med,
  exposure_label = "Metabolic\nrisk",
  mediator_label = "Inflammation",
  outcome_label = "Incident\noutcome",
  language = "en",
  palette = "jama"
)
```

The note below the plot is not decoration. It anchors three things that should
never be left implicit:

- which exposure contrast was evaluated
- which covariate profile was used
- whether the proportion mediated was automatically flagged as unstable

# Compare the built-in journal palettes

The plotting helper keeps the geometry fixed while letting you switch the
journal-style palette. That makes it easier to choose a style without
rebuilding the figure by hand.

```{r, fig.width = 10, fig.height = 8}
old_par <- graphics::par(no.readonly = TRUE)
on.exit(graphics::par(old_par), add = TRUE)
graphics::par(mfrow = c(2, 2), mar = c(1.2, 1.2, 2.2, 1.2))
for (pal in c("jama", "jco", "lancet", "nejm")) {
  leo_cox_mediation_plot(
    x = res_med,
    exposure_label = "Metabolic\nrisk",
    mediator_label = "Inflammation",
    outcome_label = "Incident\noutcome",
    language = "en",
    palette = pal,
    add_note = FALSE
  )
  graphics::title(main = pal, line = 0.2, cex.main = 1)
}
```

# What Step 3.0 should leave you with

For the default `method = "mediation"` route, the practical reading order is:

1. confirm that the engine is really `mediation` and the outcome model is
   really `survreg`
2. read ACME, ADE, ATE, and PM on the scale that is actually reported
3. check the covariate profile used for evaluation
4. treat PM cautiously if the total-effect interval crosses the null
5. switch to Step 3.1 only when you genuinely need the more explicit
   `regmedint` decomposition family

That last point is worth being strict about. Many analyses do not need the
advanced route. They need a clean question, a transparent default engine, and a
calm interpretation that stays on the scale the model actually used.
