---
title: "Survival Analysis Step 3: Mediation Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survival Analysis Step 3: Mediation Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
suppressPackageStartupMessages(library(leo.ukb))
if (!requireNamespace("regmedint", quietly = TRUE)) stop("Package 'regmedint' is required to build this tutorial.")
```

# What this step is for

This tutorial is the third survival-analysis step in `leo.ukb`. After the main
Cox model and the subgroup/interaction stage, a different question can appear:

> does part of the exposure-outcome association operate through a mediator?

That is a mediation question, not a subgroup question.

# First separate mediation from interaction

The three steps in the current survival workflow answer three different
questions:

1. Step 1: is there a main exposure-outcome association?
2. Step 2: does that association differ across groups?
3. Step 3: is part of that association transmitted through a mediator?

This distinction matters because mediation is about **pathways**, whereas
interaction is about **effect heterogeneity**.

# Know the method boundary before fitting

`leo_cox_mediation()` wraps `regmedint::regmedint()` with `yreg = "survCox"`.
The current package workflow is most appropriate when:

- the outcome is time-to-event
- the exposure is binary, or numeric with explicit `a0` and `a1`
- the mediator is binary or continuous
- the event is reasonably rare
- the temporal ordering is scientifically defensible

Two practical cautions are especially important in UK Biobank-style cohort work:

1. `regmedint` with `survCox` relies on the rare-event approximation
2. if exposure and mediator are both measured at the same baseline visit, the
   result is easier to defend as a pathway-consistent decomposition than as a
   strong causal claim unless temporal ordering is otherwise justified

# Build a compact teaching data set

We use a small synthetic example with a rare event, a binary exposure, a binary
mediator, and a few covariates.

```{r}
set.seed(20260323)
n <- 500
age <- rnorm(n, 60, 8)
bmi <- rnorm(n, 26, 4)
smoking <- factor(sample(c("Never", "Ever"), n, replace = TRUE))
exposure_num <- rbinom(n, 1, plogis(-0.2 + 0.02 * (age - 60)))
mediator_num <- rbinom(n, 1, plogis(-0.5 + 0.9 * exposure_num + 0.02 * (bmi - 26)))
event_prob <- plogis(-3.1 + 0.55 * exposure_num + 0.40 * mediator_num + 0.02 * (age - 60))

med_df <- data.frame(
  outcome = rbinom(n, 1, event_prob),
  outcome_censor = rexp(n, rate = 0.08) + 0.2,
  exposure = factor(exposure_num, levels = c(0, 1), labels = c("No", "Yes")),
  mediator = factor(mediator_num, levels = c(0, 1), labels = c("Low", "High")),
  age = age,
  bmi = bmi,
  smoking = smoking
)

mean(med_df$outcome)
```

The observed event proportion is low enough for this toy example to be aligned
with the intended `survCox` use case.

# Fit the simplest mediation workflow

As in the Cox tutorials, it helps to define the adjustment hierarchy first.

```{r}
med_models <- list(
  "Model A" = c("age"),
  "Model B" = c("age", "bmi")
)

res_med <- leo_cox_mediation(
  df = med_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "exposure",
  x_med = "mediator",
  x_cov = med_models,
  verbose = FALSE
)

res_med$result
```

# How to read the result table

The mediation output is already organized in a paper-friendly way.

- `Effect code` gives the standard mediation abbreviations such as `cde`,
  `pnde`, `tnie`, and `pm`
- `Effect` gives the full label
- `Scale` reminds you that `pm` is a proportion, while the other mediation
  effects are reported on the hazard-ratio scale
- `Exposure contrast` and `Mediator reference` show exactly where the effects
  were evaluated

This last point is important: mediation estimates are conditional on the chosen
evaluation settings.

# Inspect the evaluation settings directly

`leo_cox_mediation()` now keeps a compact evaluation summary in `$evaluation`.

```{r}
res_med$evaluation
```

This table is especially useful when you want to audit:

- which `a0 -> a1` contrast was used
- which mediator reference value was used for the controlled direct effect
- which covariate profile was used for `c_cond`

# Non-numeric covariates need explicit `c_cond`

When the selected adjustment set contains a non-numeric covariate, the
evaluation profile should be given explicitly.

```{r}
res_med_smoking <- leo_cox_mediation(
  df = med_df,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "exposure",
  x_med = "mediator",
  x_cov = "smoking",
  c_cond = list(smoking = "Never"),
  verbose = FALSE
)

res_med_smoking$evaluation
```

This is more transparent than letting a factor be silently converted in a way
the reader cannot easily reconstruct.

# Continuous mediators are also supported

The same workflow can be used with a continuous mediator.

```{r}
med_df_cont <- dplyr::mutate(
  med_df,
  mediator_cont = rnorm(n, mean = 0.5 * as.integer(exposure == "Yes") + 0.02 * (age - 60), sd = 1)
)

res_med_cont <- leo_cox_mediation(
  df = med_df_cont,
  y_out = c("outcome", "outcome_censor"),
  x_exp = "exposure",
  x_med = "mediator_cont",
  x_cov = "age",
  mediator_model = "linear",
  verbose = FALSE
)

res_med_cont$result
```

For a continuous mediator, the reference value in the table is numeric because
the controlled direct effect is evaluated at `m_cde`.

# How to write this in a paper

A defensible reporting order is usually:

1. report the main exposure-outcome association first
2. then present the mediation analysis as a pathway decomposition
3. state the exposure contrast, mediator reference, and adjustment profile
4. mention that `survCox` mediation relies on the rare-event approximation
5. avoid over-claiming causality when exposure and mediator are measured at the
   same baseline visit

In other words, mediation should usually refine the biological story after the
main Cox result is already established.

# What to take away from step 3

This step should leave you with four practical rules:

1. mediation asks a pathway question, not a heterogeneity question
2. `a0`, `a1`, `m_cde`, and `c_cond` define where the mediation effects are
   evaluated
3. `leo_cox_mediation()` handles both binary and continuous mediators, but
   factor mediators should not be forced into `mediator_model = "linear"`
4. the `survCox` mediation result should always be interpreted with the
   rare-event assumption and the study's temporal ordering in mind
